---
title: "Advancement meeting code"
editor_options: 
  chunk_output_type: console
---

```{r}
library(Seurat)
library(tidyverse)

source("03-scripts/R/seq_functions.R")
```

```{r}
nuclei <- LoadDataset("May2024")
```

# 6h analysis

Let's first perform DGE among active vs inactive cells compared to:

-   SE

-   Each other

```{r}
IEG_symbols <- c(
  'Arc',
  'Btg2',
  'Dusp1',
  'Dusp5',
  'Egr1',
  'Egr2',
  'Egr3',
  'Egr4',
  'Fos',
  'Fosb',
  'Fosl2',
  'Junb',
  'Npas4',
  'Nr4a1',
  'Nr4a3'
)

# get active and inactive cells
active_cells <- FindActiveCells(nuclei, IEG_symbols, 3)
df_active <- active_cells$df_active_cells |> 
  mutate(cell = str_sub(cell, 1, 8))

# put df_active info in Seurat metadata
meta <- nuclei@meta.data |> 
  mutate(cell = str_sub(bc_wells, 1, 8)) |> 
  relocate(cell)
meta <- meta |> 
  left_join(df_active)

nuclei <- AddMetaData(nuclei, meta)
sixhr <- nuclei |> 
  subset(subclass_name == '016 CA1-ProS Glut') |> 
  subset(condition == 'd6h')
Idents(sixhr) <- sixhr$active_binary

deg_within6h <- FindMarkers(
  sixhr, 
  logfc.threshold = 0.25,
    test.use = 'MAST',
    ident.1 = TRUE, ident.2 = FALSE,
    latent.vars = c('percent.mt'),
  ) |> 
    filter(p_val_adj < 0.05) |> 
    rownames_to_column(var = 'gene')
```

Sweet. Now let's find the DEGs between 6h vs SE for both active and inactive cells

```{r}
barcodes_6h_active <- nuclei@meta.data |> 
  filter(subclass_name == '016 CA1-ProS Glut') |> 
  filter(condition == 'd6h') |> 
  filter(active_binary == TRUE) |> 
  pull(cell) |> print()

barcodes_6h_inactive <- nuclei@meta.data |> 
  filter(subclass_name == '016 CA1-ProS Glut') |> 
  filter(condition == 'd6h') |> 
  filter(active_binary == FALSE) |> 
  pull(cell)

barcodes_missing_inactive <- nuclei@meta.data |> 
  filter(!cell %in% barcodes_6h_inactive) |> 
  pull(cell)

barcodes_missing_active <- nuclei@meta.data |> 
  filter(!cell %in% barcodes_6h_active) |> 
  pull(cell)

nuclei_missing_active <- nuclei |> 
  subset(cell %in% barcodes_missing_active) |> 
  subset(subclass_name == '016 CA1-ProS Glut')
nuclei_missing_inactive <- nuclei |> 
  subset(cell %in% barcodes_missing_inactive) |> 
  subset(subclass_name == '016 CA1-ProS Glut')

Idents(nuclei_missing_active) <- nuclei_missing_active$condition
Idents(nuclei_missing_inactive) <- nuclei_missing_inactive$condition

DEGs_inactive_vs_SE <- FindMarkers(
  nuclei_missing_active, 
  logfc.threshold = 0.25,
    test.use = 'MAST',
    ident.1 = 'd6h', ident.2 = 'dSE',
    latent.vars = c('percent.mt'),
  ) |> 
    filter(p_val_adj < 0.05) |> 
    rownames_to_column(var = 'gene')

DEGs_active_vs_SE <- FindMarkers(
  nuclei_missing_inactive, 
  logfc.threshold = 0.25,
    test.use = 'MAST',
    ident.1 = 'd6h', ident.2 = 'dSE',
    latent.vars = c('percent.mt'),
  ) |> 
    filter(p_val_adj < 0.05) |> 
    rownames_to_column(var = 'gene')
```

Let's do some GSEA for these genes

```{r}
library(clusterProfiler)
library(org.Mm.eg.db)

# turn gene and avg_log2FC into a named vector
gene_list <- DEGs_active_vs_SE$avg_log2FC
names(gene_list) <- DEGs_active_vs_SE$gene
gene_list <- sort(gene_list, decreasing = TRUE)

gso_inactive_vs_SE <- enrichGO(
  gene     = DEGs_inactive_vs_SE$gene,
  OrgDb    = org.Mm.eg.db,
  keyType = "SYMBOL",
  ont      = "CC"
)

gso_active_vs_SE <- enrichGO(
  gene     = DEGs_active_vs_SE$gene,
  pvalueCutoff = 1,
  OrgDb    = org.Mm.eg.db,
  keyType = "SYMBOL",
  ont      = "CC"
)
df <- gso_active_vs_SE@result
```

```{r}
performGSEA <- function(dge_df, contrast_name) {
  library(clusterProfiler)
  library(org.Mm.eg.db)
  library(enrichplot)
  
  # do GO analysis
  go_results <- enrichGO(
  gene     = dge_df$gene,
  pvalueCutoff = 1,
  qvalueCutoff = 1,
  OrgDb    = org.Mm.eg.db,
  keyType = "SYMBOL",
  ont      = "CC"
  )
  
  df <- go_results@result
  df$contrast <- contrast_name
  
  return(df)
}
```

# plotting GO terms across several conditions

To create a plot matching the carpet-heatmap format of DGE for all classes, I should make a geom_tile with X being the conditions and y being the GO terms themselves. I want to color the tiles by z-score of the GO term. I suppose this could also be a geom_point.

**Pseudocode:**

1.  Load in DEGs for all contrasts involving dSE.
2.  Perform enrichGO on all contrasts.
3.  Assemble all GO terms into one df, denoting which contrast they came from.
4.  Create a geom_point. Aes:
    1.  x = contrast
    2.  y = GO term
    3.  size = zScore
    4.  color = p.adjust

```{r steps 1-3}
DEGs_30m_vs_SE <- read_csv("04-analysis/DEGs/condition/016_CA1-ProS_Glut_d30m_vs_dSE.csv")
DEGs_6h_vs_SE <- read_csv("04-analysis/DEGs/condition/016_CA1-ProS_Glut_d6h_vs_dSE.csv")
DEGs_KA_vs_SE <- read_csv("04-analysis/DEGs/condition/016_CA1-ProS_Glut_KA_vs_dSE.csv")

results_30m_vs_SE <- performGSEA(DEGs_30m_vs_SE, "d30m")
results_6h_vs_SE <- performGSEA(DEGs_6h_vs_SE, "d6h")
results_KA_vs_SE <- performGSEA(DEGs_KA_vs_SE, "KA")

df_go_terms <- bind_rows(results_30m_vs_SE, results_6h_vs_SE, results_KA_vs_SE)

significant_terms <- df_go_terms |> 
  filter((contrast == "d30m" & qvalue < 0.05) |
         (contrast == "d6h" & qvalue < 0.05))  |> 
  pull(ID)

term_levels <- df_go_terms |> 
  filter(ID %in% significant_terms) |> 
  filter(contrast != 'KA') |> 
  group_by(ID) |> 
  slice_max(order_by = zScore, n = 1)  |> 
  ungroup()  |> 
  arrange(contrast) |> 
  pull(ID)
  
  
# add levels to df_go_terms
df_go_terms <- df_go_terms |> 
  filter(ID %in% term_levels) |> 
  print()
```

```{r plotting}
ggplot(df_go_terms, aes(x = contrast, y = Description)) +
  geom_tile(aes(fill = zScore)) +
  scale_color_gradient(low = "blue", high = "red", name = "-log10(q)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Contrast", y = "GO Term",
       size = "Count", # Change this to z-score if available
       title = "GO Term Enrichment Across Contrasts")
```

# co-expression

```{r}
# filter nuclei for each condition, and then within condition pick out active cells
AddMetaData(nuclei, df_active_cells_IEGs)
ca1 <- nuclei |> subset(subclass_name == '016 CA1-ProS Glut')

# add binary metadata
meta <- ca1@meta.data
meta <- meta |> 
  left_join(df_active_cells_IEGs, by = c('bc_wells' = 'cell'))
ca1 <- AddMetaData(ca1, meta)
ca1_active <- ca1 |> subset(active_binary == TRUE)
```


```{r}
corr_SE_all <- GetCorrData(ca1, 'dSE', IEG_symbols)
corr_SE_active <- GetCorrData(ca1_active, 'dSE', IEG_symbols)
PlotComplexHeatmap(corr_SE_all, plot_title = 'dSE all')
PlotComplexHeatmap(corr_SE_active, plot_title = 'dSE active')

corr_30m_all <- GetCorrData(ca1, 'd30m', IEG_symbols)
corr_30m_active <- GetCorrData(ca1_active, 'd30m', IEG_symbols)
PlotComplexHeatmap(corr_30m, plot_title = 'd30m all')
PlotComplexHeatmap(corr_30m_active, plot_title = 'd30m active')

corr_6h_all <- GetCorrData(ca1, 'd6h', IEG_symbols)
corr_6h_active <- GetCorrData(ca1_active, 'd6h', IEG_symbols)
PlotComplexHeatmap(corr_6h_all, plot_title = 'd6h all')
PlotComplexHeatmap(corr_6h_active, plot_title = 'd6h active')

corr_KA_all <- GetCorrData(ca1, 'KA', IEG_symbols)
corr_KA_active <- GetCorrData(ca1_active, 'KA', IEG_symbols)
PlotComplexHeatmap(corr_KA_all, plot_title = 'KA all')
PlotComplexHeatmap(corr_KA_active, plot_title = 'KA active')
```
