---
title: "Merge all seq data"
author: "jack olmstead"
format: html
editor_options: 
  chunk_output_type: console
---

I want to merge all my sequencing data into a single Seurat object and perform all the same processing steps on it.

```{r}
library(tidyverse)
library(Seurat)
library(plotly)
library(readxl)
```

```{r create Seurats}
# get main folders for each experiment
dir_merge <- "/Volumes/jack/seq/analysis_merge/combined/0_all-sample/DGE_filtered/"
dir_May2024 <- "/Volumes/jack/seq/analysis_May2024/combined/0_all-sample/DGE_filtered/"

# read in matricies from each seq experiment
mat_merge <- ReadParseBio(dir_merge)
mat_May2024 <- ReadParseBio(dir_May2024)

# make Seurat objects for each
nuclei_merge <- CreateSeuratObject(mat_merge)
nuclei_May2024 <- CreateSeuratObject(mat_May2024)

# add new columns to seurat objects
nuclei_merge$experiment <- 'merge'
nuclei_May2024$experiment <- 'May2024'

nuclei_merge$zt <- 4
nuclei_May2024$zt <- 12

nuclei_merge
nuclei_May2024
```

```{r ReadMeta()}
# define a function to read in metadata for each dataset
ReadMeta <- function(data_dir, seurat_obj){
  cell_meta <- read.delim(paste0(data_dir, "cell_metadata.csv"),
                      stringsAsFactor = FALSE, sep = ",")

  # for allen metadata, find the csv inside a subfolder that starts with 'adata_lean'
  allen_meta_folder <- list.files(data_dir, pattern = 'adata_lean', full.names = TRUE)
  allen_meta_folder <- allen_meta_folder[file.info(allen_meta_folder)$isdir]
  allen_meta_csv <- list.files(allen_meta_folder, pattern = '.csv', full.names = TRUE)
  
  allen_meta <- read.csv(allen_meta_csv, stringsAsFactor = FALSE, sep = ",", skip = 4)
  
  meta <- merge(cell_meta, allen_meta, by.x = 'bc_wells', by.y = 'cell_id')
  
  # add condition and sublibrary metadata
  meta <- meta |> 
    mutate(condition = substr(sample, 1, nchar(sample)-1)) |> 
    mutate(sublibrary = factor(substr(bc_wells, nchar(bc_wells), nchar(bc_wells))))
  
  # rename condition from 'dSE' to 'SE' for consistency
  if (seurat_obj$experiment[1] == 'May2024') {
    meta$condition <- gsub('dSE', 'SE', meta$condition)
    meta$condition <- gsub('d6h', '6h', meta$condition)
    meta$condition <- gsub('d30m', '30m', meta$condition)
    # relevel condition
    meta$condition <- factor(meta$condition, levels = c('SE', '30m', '6h', 'KA'))
  } else {
    meta$condition <- factor(meta$condition, levels = c('SE', '1h', '6h'))
  }
  
  # reorder metadata according to Seurat object
  seurat_obj <- subset(seurat_obj, cells = meta$bc_wells)
  meta <- meta |> arrange(match(bc_wells, rownames(seurat_obj@meta.data)))
  
  # add metadata to seurat object
  seurat_obj_with_meta <- seurat_obj |> AddMetaData(meta)
  
  return(seurat_obj_with_meta)
}
```

```{r add metadata to Seurats}
nuclei_merge <- ReadMeta(dir_merge, nuclei_merge)
nuclei_May2024 <- ReadMeta(dir_May2024, nuclei_May2024)

nuclei_merge
nuclei_May2024
```

```{r normalize Seurats}
# calculate percent mt for each dataset
nuclei_merge$percent.mt <- PercentageFeatureSet(nuclei_merge, pattern = "^mt-")
nuclei_May2024$percent.mt <- PercentageFeatureSet(nuclei_May2024, pattern = "^mt-")

# normalize data
nuclei_merge <- SCTransform(nuclei_merge, vars.to.regress = 'percent.mt')
nuclei_May2024 <- SCTransform(nuclei_May2024, vars.to.regress = 'percent.mt')
```

```{r merge Seurats}
# merge all seurat objects
nuclei <- merge(nuclei_merge, nuclei_May2024, 
                add.cell.ids = c('merge','May2024'),
                merge.data = TRUE
                )
```

```{r relevel metadata in merged Seurat}
# relevel condition
nuclei$condition <- factor(nuclei$condition, levels = c('SE', '30m', '1h', '6h', 'KA'))
nuclei$zt <- factor(nuclei$zt, levels = c(4, 12))
```

```{r PCA and UMAP}
# get variable features for the merged dataset via the intersection of vfs from each dataset
vf_merge <- as_tibble(VariableFeatures(nuclei_merge))
vf_May2024 <- as_tibble(VariableFeatures(nuclei_May2024))
vf <- inner_join(vf_merge, vf_May2024)
VariableFeatures(nuclei) <- vf$value

nuclei <- nuclei |> 
  RunPCA() |>
  FindNeighbors(dims = 1:41) |>
  FindClusters(resolution = 0.8) |>
  BuildClusterTree(reorder = TRUE, reorder.numeric = "TRUE") |>
  RunUMAP(dims = 1:41, n.components = 3)
```

```{r 3-D UMAP}
# nuclei |> 
#   plot_ly(
#     x = ~`umap_1`,
#     y = ~`umap_2`,
#     z = ~`umap_3`,
#     color = ~zt
#   )
```

```{r}
nuclei <- PrepSCTFindMarkers(nuclei, verbose = T)
```

```{r}
deg_zt4 <- nuclei |> 
  subset(subclass_name == '016 CA1-ProS Glut') |> 
  subset(zt == '4') |> 
  FindMarkers(group.by = 'condition', ident.1 = '1h', ident.2 = 'SE') |> 
  rownames_to_column(var = 'gene') |>
  # filter(p_val_adj < 0.05) |> 
  # filter(abs(avg_log2FC) > 0.5) |> 
  arrange(desc(avg_log2FC))

deg_zt12 <- nuclei |>
  subset(subclass_name == '016 CA1-ProS Glut') |>
  subset(zt == '12') |>
  FindMarkers(group.by = 'condition', ident.1 = '30m', ident.2 = 'SE') |>
  rownames_to_column(var = 'gene') |>
  # filter(p_val_adj < 0.05) |>
  # filter(abs(avg_log2FC) > 0.5) |>
  arrange(desc(avg_log2FC))

deg <- left_join(deg_zt4, deg_zt12, by = 'gene', suffix = c('_zt4', '_zt12'))
```

```{r define IEGs}
iegs <- c(
'Egr2',
'Homer1',
'Egr3',
'Map2',
'Dusp5',
'Per2',
'Dusp6',
'Per1',
'Arc',
'Rhoq'
)
```

```{r define seizure genes}
# read in seizure genes
library(readxl)
seizure_genes <- read_excel("/Volumes/jack/seq/analysis_May2024/combined/0_all-sample/DGE_filtered/DEGs/SCT_normalized/016 CA1-ProS Glut_dSE_KA_SIGNIF.xlsx")
```

```{r plot ZT4 vs ZT12}
p1 <- deg |>
  filter(!(gene %in% seizure_genes$gene)) |> 
  ggplot() +
  geom_point(aes(x = pct.2_zt4, y = pct.2_zt12)) +
  geom_abline(slope = 1) +
  coord_fixed(ratio = 1) +
  labs(title = 'Ratio of percent of SE cells expressing IEGs',
       x = 'ZT4',
       y = 'ZT12')

p2 <- deg |>
  filter(!(gene %in% seizure_genes$gene)) |> 
  ggplot() +
  geom_point(aes(x = pct.1_zt4, y = pct.1_zt12)) +
  geom_abline(slope = 1) +
  coord_fixed(ratio = 1) +
  labs(title = 'Ratio of percent of EE cells expressing IEGs',
       x = 'ZT4',
       y = 'ZT12')

p3 <- deg |> 
  filter(!(gene %in% seizure_genes$gene)) |> 
  ggplot() +
  geom_point(aes(x = avg_log2FC_zt4, y = avg_log2FC_zt12)) +
  coord_fixed(xlim = c(-3,3), ylim = c(-3,3)) +
  geom_abline(slope = 1) +
  labs(title = 'Ratio of log2FC between SE vs EE',
       x = 'ZT4',
       y = 'ZT12')

p1 + p2
```

Alright, this is beginning to be concerning. I'm  seeing that almost all genes are systematically upregulated in ZT12 compared to ZT4. This pattern holds for genes that are selected to *not* be in seizure genes, i.e. those that we would not expect to be related to activity. This is a large blow against the hypothesis that IEGs are expressed in a circadian manner.

Another note: I should separately just calculate the DEGs between ZT4 and ZT12 for SE and EE, respectively.

```{r}
deg_se <- nuclei |>
  subset(subclass_name == '016 CA1-ProS Glut') |>
  subset(condition == 'SE') |> 
  FindMarkers(group.by = 'zt', ident.1 = '4', ident.2 = '12', recorrect_umi = F) |>
  rownames_to_column(var = 'gene') |>
  arrange(desc(avg_log2FC))
  
deg_ee <- nuclei |>
  subset(subclass_name == '016 CA1-ProS Glut') |>
  subset(condition == '30m' | condition == '1h') |> 
  FindMarkers(group.by = 'zt', ident.1 = '4', ident.2 = '12', recorrect_umi = F) |>
  rownames_to_column(var = 'gene') |>
  arrange(desc(avg_log2FC))
```

```{r}
p1 <- deg_se |>
  ggplot() +
  geom_point(aes(x = pct.1, y = pct.2)) +
  geom_abline(slope = 1, color = 'red') +
  coord_fixed(ratio = 1) +
  labs(title = 'Ratio of percent expression by ZT in SE conditions',
       x = 'ZT4',
       y = 'ZT12')

p2 <- deg_ee |>
  ggplot() +
  geom_point(aes(x = pct.1, y = pct.2)) +
  geom_abline(slope = 1, color = 'red') +
  coord_fixed(ratio = 1) +
  labs(title = 'Ratio of percent expression by ZT in EE conditions',
       x = 'ZT4',
       y = 'ZT12')

p1 + p2
```

```{r}
p1 <- plot_ly(
  data = deg_se,
  x = ~pct.1,
  y = ~pct.2,
  type = 'scatter',
  mode = 'markers',
  marker = list(opacity = 0.3),
  text = ~gene
) |> 
  add_segments(x = 0, xend = 1, y = 0, yend = 1, line = list(color = 'red'))

p2 <- plot_ly(
  data = deg_ee,
  x = ~pct.1,
  y = ~pct.2,
  type = 'scatter',
  mode = 'markers',
  marker = list(opacity = 0.3),
  text = ~gene
) |> 
  add_segments(x = 0, xend = 1, y = 0, yend = 1, line = list(color = 'red'))

subplot(p1, p2) |> 
  layout(
    xaxis = list(title = 'ZT4'),
    yaxis = list(title = 'ZT12'),
    title = 'Ratio of percent expression in SE and EE conditions by ZT'
  )
```

