---
title: "Compare flattop Seurat"
---

Compare flattop Seurat to normal Seurat with flattop reads still in there. Perform the following specific analyses:

1.  Side-by-side comparison of UMAPs

2.  Basic comparisons:

    1.  Scatterplots:

        1.  Transcript count filtered vs. transcript count unfiltered

        2.  Gene count filtered vs. gene count unfiltered

```{r}
#| message: false

library(Seurat)
library(ggplot2)
library(readr)
library(dplyr)
library(glue)
library(patchwork)

# source("seq_functions.R")
source("03-scripts/R/seq_functions.R")

condition_colors <- LoadConditionColors("May2024")
subclass_colors <- LoadAllenColors()
```

# Data prep

The Seurat object for the library with all flattop reads filtered out will be called `nuclei_filtered`, and the other one will be called `nuclei_unfiltered`.

```{r load Seurat objects}
nuclei_filtered <- LoadSeuratRds("~/projects/flattop/02-data/raw_data/cutoff80_parse_outputs/all-sample/DGE_filtered/seurat.Rds")
nuclei_unfiltered <- LoadDataset("May2024")
```

```{r merge metadatas}
# take the sublibrary tag off of barcodes for unfiltered lib
nuclei_unfiltered$bc_wells <- str_sub(nuclei_unfiltered$bc_wells, 1, 8)

# filtered nuclei have x columns, unfiltered have x columns
meta <- nuclei_unfiltered@meta.data |> 
  left_join(nuclei_filtered@meta.data, by = "bc_wells")
```

There are many cells in nuclei_unfiltered that got removed from nuclei_filtered. Let's make a Seurat for these cells specifically.

```{r nuclei_lost}
lost_cell_bc <- meta |> 
  filter(is.na(gene_count.y)) |> 
  pull(bc_wells)

nuclei_lost <- subset(nuclei_unfiltered, bc_wells %in% lost_cell_bc)
nuclei_lost
```

Remove all lost cells from meta, since it doesn't have info for both filtered and unfiltered

```{r}
meta <- meta |> 
  filter(!bc_wells %in% nuclei_lost$bc_wells)

meta_s1 <- meta |> filter(sublibrary == 1)
meta_s2 <- meta |> filter(sublibrary == 2)
```

# UMAPs

```{r}
umap_theme <- theme(
  panel.grid = element_blank(),
    axis.text = element_blank(),
    legend.text = element_text(size = 12),
    legend.position = 'none',
  panel.border = element_rect(fill=NA),
  plot.title = element_text(size = 25)
)
# condition plots
p1 <- DimPlot(nuclei_unfiltered, reduction = "umap", group.by = "condition", pt.size = 1, shuffle = TRUE, seed = 17) +
  scale_color_manual(values = condition_colors) +
  theme_minimal() +
  labs(title = 'Unfiltered', x = '', y = '') +
  umap_theme

p2 <- DimPlot(nuclei_filtered, reduction = "umap", group.by = "condition", pt.size = 1, shuffle = TRUE, seed = 17) +
  scale_color_manual(values = condition_colors) +
  theme_minimal() +
  labs(title = 'Filtered', x = '', y = '') +
  umap_theme

# subclass plots
p3 <- DimPlot(nuclei_unfiltered, reduction = "umap", group.by = "subclass_name", pt.size = 1, shuffle = TRUE, seed = 17) +
  scale_color_manual(values = subclass_colors) +
  theme_minimal() +
  labs(title = 'Unfiltered', x = '', y = '') +
  umap_theme

p4 <- DimPlot(nuclei_filtered, reduction = "umap", group.by = "subclass_name", pt.size = 1, shuffle = TRUE, seed = 17) +
  scale_color_manual(values = subclass_colors) +
  theme_minimal() +
  labs(title = 'Filtered', x = '', y = '') +
  umap_theme

(p1 + p2) / (p3 + p4)
```

# Transcript/gene correlation

```{r}
corr_theme <- theme(
  legend.position = 'none',
  plot.title = element_text(size = 20),
  axis.title.x = element_text(size = 20),
  axis.title.y = element_text(size = 20),
  axis.text.x = element_text(size = 15),
  axis.text.y = element_text(size = 15),
)

plot_transcripts <- ggplot(meta) +
  aes(x = tscp_count.x, y = tscp_count.y) +
  geom_density_2d_filled() +
  geom_point(color = 'white', size = 3, alpha = 0.5) +
  theme_minimal() +
  labs(
    title = 'Transcripts',
    x = 'Before',
    y = 'After',
  ) +
  corr_theme

plot_genes <- ggplot(meta) +
  aes(x = gene_count.x, y = gene_count.y) +
  geom_density_2d_filled() +
  geom_point(color = 'white', size = 3, alpha = 0.5) +
  theme_minimal() +
  labs(
    title = 'Genes',
    x = 'Before',
    y = 'After',
  ) +
  corr_theme

plot_transcripts + plot_genes + plot_annotation(
  title = "Comparison of transcripts and genes per cell before/after flattop read removal",
  caption = "Each point is a cell, plotted as number of detected transcripts/reads after vs. before flattop read removal.",
  theme = theme(
    plot.title = element_text(size = 30),
    plot.caption = element_text(size = 15)
    )
  )
```

Make the exact same plot; just color the points by sublibrary.

```{r sublib coloring}
corr_theme <- theme(
  legend.position = 'none',
  plot.title = element_text(size = 20),
  axis.title.x = element_text(size = 20),
  axis.title.y = element_text(size = 20),
  axis.text.x = element_text(size = 15),
  axis.text.y = element_text(size = 15),
)
sublib_colors <- setNames(c("white", "red"), c(1,2))

plot_transcripts <- ggplot(meta) +
  aes(x = tscp_count.x, y = tscp_count.y) +
  geom_density_2d_filled() +
  geom_point(aes(color = sublibrary), size = 3, alpha = 0.5) +
  scale_color_manual(values = sublib_colors) +
  theme_minimal() +
  labs(
    title = 'Transcripts',
    x = 'Before',
    y = 'After',
  ) +
  corr_theme

plot_genes <- ggplot(meta) +
  aes(x = gene_count.x, y = gene_count.y) +
  geom_density_2d_filled() +
  geom_point(aes(color = sublibrary), size = 3, alpha = 0.5) +
  scale_color_manual(values = sublib_colors) +
  theme_minimal() +
  labs(
    title = 'Genes',
    x = 'Before',
    y = 'After',
  ) +
  corr_theme

plot_transcripts + plot_genes + plot_annotation(
  title = "Comparison of transcripts and genes per cell before/after flattop read removal",
  caption = "Each point is a cell, plotted as number of detected transcripts/reads after vs. before flattop read removal. Red points = sublibrary 2; White points = sublibrary 1.",
  theme = theme(
    plot.title = element_text(size = 30),
    plot.caption = element_text(size = 15)
    )
  )
```

# Correlations

Let's find the correlations between detected transcript and gene count before/after filtering out flattop reads.

```{r}
# both sublibraries R2
r2_tscp <- cor(meta$tscp_count.x, meta$tscp_count.y)^2
r2_gene <- cor(meta$gene_count.x, meta$gene_count.y)^2

print(glue("Combined sublibrary transcript count correlation R-sq: {r2_tscp}"))
print(glue("Combined sublibrary gene count correlation R-sq: {r2_gene}"))

# sublib 1 R2
r2_tscp <- cor(meta_s1$tscp_count.x, meta_s1$tscp_count.y)^2
r2_gene <- cor(meta_s1$gene_count.x, meta_s1$gene_count.y)^2

print(glue("Sublibrary 1 transcript count correlation R-sq: {r2_tscp}"))
print(glue("Sublibrary 1 gene count correlation R-sq: {r2_gene}"))

# sublib 2 R2
r2_tscp <- cor(meta_s2$tscp_count.x, meta_s2$tscp_count.y)^2
r2_gene <- cor(meta_s2$gene_count.x, meta_s2$gene_count.y)^2

print(glue("Sublibrary 1 transcript count correlation R-sq: {r2_tscp}"))
print(glue("Sublibrary 1 gene count correlation R-sq: {r2_gene}"))

# sublib distributions
print("original sublib distribution:")
table(nuclei_unfiltered$sublibrary)

print("filtered sublib distribution:")
table(meta$sublibrary)
```
