---
title: "Plotting IEG expression"
---

# Prep data

```{r setup}
library(tidyverse)
library(ggridges)
library(Seurat)
library(ggh4x)

source("03-scripts/R/seq_functions.R")
```

```{r setup}
nuclei <- LoadDataset("May2024")
activity_colors <- LoadActivityColors("May2024")
```

```{r}
LRG_symbols <- c(
  'Arc',
  'Btg2',
  'Dusp1',
  'Dusp5',
  'Egr1',
  'Egr2',
  'Egr3',
  'Egr4',
  'Fos',
  'Fosb',
  'Fosl2',
  'Junb',
  'Npas4',
  'Nr4a1',
  'Nr4a3'
)

# LRG_symbols <- c(
#   "Fosl2",
#   "Gpr3",
#   "Rgs2",
#   "Vgf",
#   "Adcyap1",
#   "Bdnf",
#   "Nrn1",
#   "Ptgs2",
#   "Nptx2",
#   "Homer1"
# )

# up-regulated LRGs
LRG_symbols <- read_csv("04-analysis/DEGs/condition/016_CA1-ProS_Glut_d6h_vs_SE.csv") |> 
  filter(avg_log2FC > 0) |> 
  filter(!gene %in% c("4732440D04Rik", "9330162G02Rik", "5031425E22Rik", "Cyp2c52-ps", "0610043K17Rik", "2210408F21Rik", "Mir124a-1hg")) |> 
  pull(gene)

# down-regulated LRGs
LRG_symbols <- read_csv("04-analysis/DEGs/condition/016_CA1-ProS_Glut_d6h_vs_SE.csv") |> 
  filter(avg_log2FC < 0) |> 
  filter(!gene %in% c("4732440D04Rik", "9330162G02Rik", "5031425E22Rik", "Cyp2c52-ps", "0610043K17Rik", "2210408F21Rik", "Mir124a-1hg", "1700057H15Rik", "2210408I21Rik", "Rpl29-ps2", "4930419G24Rik")) |> 
  pull(gene)

# find any genes missing from CA1 for some reason
missing_columns <- setdiff(LRG_symbols, colnames(mati_ca1))
print(missing_columns)
```

```{r mati}
# get gene names for these cells
gene_names <- rownames( nuclei@assays$SCT@counts)
all(LRG_symbols %in% gene_names)

mat <- nuclei@assays$SCT@counts
mati <- t(mat[LRG_symbols, ])
meta <- nuclei@meta.data
```

```{r get count data}
# convert mati to a dataframe
mati_df <- data.frame(mati) |> 
  rownames_to_column("barcode") |> 
  left_join(meta, by = c("barcode" = "bc_wells"))

# change the condition labels
mati_df <- mati_df |> 
  mutate(
    condition = str_replace(condition, "SE", "SE"),
    condition = str_replace(condition, "d30m", "30m"),
    condition = str_replace(condition, "d6h", "6h"),
    condition = fct_relevel(condition, c("SE", "30m", "6h", "KA"))
  )

mati_df$sample <- factor(mati_df$sample, levels = c("SE1", "SE2", "SE3", 
                                                    "d30m1", "d30m2", "d30m3", 
                                                    "d6h1", "d6h2", "d6h3",
                                                    "KA1", "KA2", "KA3"))

mati_ca1 <- mati_df[mati_df$subclass_name == "016 CA1-ProS Glut", ]
# normalize
mati_ca1[, LRG_symbols] <- log1p(mati_ca1[, LRG_symbols])

# Gather data into long format
ca1_long <- pivot_longer(mati_ca1, cols = all_of(IEG_symbols), names_to = "gene", values_to = "expression")
```

Now we have a tibble with IEG expression information and metadata for all cells. We can use this to plot IEG expression in different conditions and sublibraries.

# CA1 neurons

Let's first plot expression for all CA1 neurons.

```{r violin}
# Plot with facet_wrap
p <- ggplot(ca1_long, aes(x = condition, y = expression, fill=condition)) +
  geom_violin() +
  facet_wrap(~gene, scales = "free_y") +
  scale_fill_manual(values = condition_colors) +
  labs(title = "IEG expression in all CA1 cells",
       x='', y='', fontsize=30)

print(p)
```

```{r boxplot}
p <- ggplot(ca1_long, aes(x = condition, y = expression, fill=condition)) +
  geom_boxplot() +
  facet_wrap(~gene, scales = "free_y") +
  scale_fill_manual(values = condition_colors) +
  labs(title = "IEG expression in all CA1 cells",
       x='', y='', fontsize=30) 

print(p)
```

## Percent cells expressing an IEG

Up to this point we've been looking at dimensionalized expression.

Let's examine binary expression now: Proportion of cells in each replicate that express an IEG.

```{r }
# make a new col ca1_long$expressed where expression > 0 is 1 and expression = 0 is 0
ca1_long$expressed <- ifelse(ca1_long$expression > 0, 1, 0)

p <- ggplot(ca1_long, aes(x = activity_condition, fill = activity_condition)) +
  geom_bar(aes(y = ifelse(expressed > 0, 1, 0)), stat = "summary", fun = "mean", color = 'black') +
  scale_fill_manual(values = activity_colors) +
  facet_wrap(~gene, scales = "free_y") +
  labs(title = "Proportion of CA1 cells expressing IEGs",
       x = '', y = 'Proportion', fontsize = 30) +
  theme(
    plot.title = element_text(size = 30),
    strip.text = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    axis.text.y = element_text(size = 15),
    axis.text.x = element_text(size = 20),
    legend.position = 'none'
  )

print(p)
```

## examining SE replicates

### all SE cells

Let's examine the SE replicates in-depth, visualizing the distribution for each sample.

```{r}
# first, filter out all rows where condition is not "SE"
ca1_long_SE <- ca1_long[ca1_long$condition == "SE", ]
```

```{r }
p <- ggplot(ca1_long_SE, aes(x = sample, y = expression)) +
  geom_boxplot(fill = "#FCF6B1") +
  facet_wrap(~gene, scales = "free_y") +
  labs(title = "CA1 IEG expression in SE replicates",
       x='', y='', fontsize=30)

print(p)
```

```{r }
p <- ggplot(ca1_long_SE, aes(x = sample, y = expression)) +
  geom_violin(fill = "#FCF6B1") +
  facet_wrap(~gene, scales = "free_y") +
  labs(title = "CA1 IEG expression in SE replicates",
       x='', y='', fontsize=30)

print(p)
```

```{r }
p <- ggplot(ca1_long_SE, aes(x = sample, y = expression)) +
  geom_jitter(shape=21, fill = "#FCF6B1", color='black') +
  facet_wrap(~gene, scales = "free_y") +
  labs(title = "CA1 IEG expression in SE replicates",
       x='', y='', fontsize=30)

print(p)
```

### only IEG-expressing SE cells

```{r}
# filter out all rows where expression is 0 in ca1_long
ca1_nonzero_SE <- ca1_long_SE[ca1_long_SE$expression > 0, ]
```

```{r }
p <- ggplot(ca1_nonzero_SE, aes(x = sample, y = expression)) +
  geom_boxplot(fill = "#FCF6B1") +
  facet_wrap(~gene, scales = "free_y") +
  labs(title = "CA1 IEG expression in SE replicates after filtering out 0s",
       x='', y='', fontsize=30)

print(p)
```

```{r }
p <- ggplot(ca1_nonzero_SE, aes(x = sample, y = expression)) +
  geom_violin(fill = "#FCF6B1") +
  facet_wrap(~gene, scales = "free_y") +
  labs(title = "CA1 IEG expression in SE replicates after filtering out 0s",
       x='', y='', fontsize=30)

print(p)
```

```{r }
p <- ggplot(ca1_nonzero_SE, aes(x = sample, y = expression)) +
  geom_jitter(shape=21, fill = "#FCF6B1", color='black') +
  facet_wrap(~gene, scales = "free_y") +
  labs(title = "CA1 IEG expression in SE replicates after filtering out 0s",
       x='', y='', fontsize=30)

print(p)
```
