---
title: "Plotting IEG expression"
editor_options: 
  chunk_output_type: console
---

# Prep data

```{r setup}
library(tidyverse)
library(ggridges)
library(Seurat)

source("03-scripts/R/seq_functions.R")
```

```{r setup}
nuclei <- LoadDataset("Dec2024_pilot")
activity_colors <- LoadActivityColors("Dec2024")
ZT_colors <- LoadZTColors(palette = 5)
```

# Gather data

```{r subset normalized expresion matrix for just IEGs}
IEG_symbols <- LoadGeneList()

# get gene names for these cells
gene_names <- rownames( nuclei@assays$RNA['data'] )
all(IEG_symbols %in% gene_names)

mat <- nuclei@assays$RNA['data']
mati <- t(mat[IEG_symbols, ])
meta <- nuclei@meta.data
```

```{r reshape matrix to df}
# convert mati to a dataframe
mati_df <- data.frame(mati) |> 
  rownames_to_column("barcode") |> 
  left_join(meta, by = c("barcode" = "bc_wells"))

mati_ca1 <- mati_df[mati_df$subclass_name == "016 CA1-ProS Glut", ]
# normalize
mati_ca1[, IEG_symbols] <- log1p(mati_ca1[, IEG_symbols])

# Gather data into long format
ca1_long <- pivot_longer(mati_ca1, cols = all_of(IEG_symbols), names_to = "gene", values_to = "expression")
```

Now we have a tibble with IEG expression information and metadata for all cells. We can use this to plot IEG expression in different conditions and sublibraries.

sh146 into sh164

ssh -x

# Plot IEG expression

```{r violin}
# Plot with facet_wrap
p <- ggplot(ca1_long, aes(x = activity_condition, y = expression, fill=activity_condition)) +
  geom_violin() +
  facet_wrap(~gene, scales = "free_y") +
  scale_fill_manual(values = activity_colors) +
  labs(title = "IEG expression in all CA1 cells",
       x='', y='', fontsize=30)

print(p)
```

```{r boxplot}
p <- ggplot(ca1_long, aes(x = activity_condition, y = expression, fill=activity_condition)) +
  geom_boxplot() +
  facet_wrap(~gene, scales = "free_y") +
  scale_fill_manual(values = activity_colors) +
  labs(title = "IEG expression in all CA1 cells",
       x='', y='', fontsize=30)

print(p)
```

## Activity-based expression

Up to this point we've been looking at dimensionalized expression.

Let's examine binary expression now: Proportion of cells in each replicate that express an IEG.

```{r by activity}
p <- ca1_long |> 
  mutate(expressed = ifelse(expression > 1, 1, 0)) |> 
ggplot(ca1_long, aes(x = activity_condition, fill = activity_condition)) +
  geom_bar(aes(y = ifelse(expressed > 0, 1, 0)), stat = "summary", fun = "mean", color = 'black') +
  scale_fill_manual(values = activity_colors) +
  facet_wrap(~gene, scales = "free_y") +
  labs(title = "Proportion of cells expressing IEGs",
       x = '', y = 'Proportion', fontsize = 30) +
  theme(
    plot.title = element_text(size = 30),
    strip.text = element_text(size = 15),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 20)
  )

print(p)
```

## ZT-based expression

```{r by ZT}
ca1_long |> 
  mutate(expressed = ifelse(expression > 1, 1, 0)) |> 
  filter(!activity_condition %in% c('KA_30m', 'KA_6h')) |>
ggplot(aes(x = ZT, fill = ZT)) +
  geom_bar(aes(y = ifelse(expressed > 0, 1, 0)), stat = "summary", fun = "mean", color = 'black') +
  scale_fill_manual(values = ZT_colors) +
  facet_grid2(vars(activity_condition), vars(gene), scales = "free", independent = 'x') +
  labs(title = "Proportion of cells expressing IEGs",
       x = '', y = 'Relative proportion', fontsize = 30) +
  theme(
    plot.title = element_text(size = 30),
    axis.title.y = element_text(size = 20),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    strip.text.y.left = element_text(),
    strip.text = element_text(size = 15),
    legend.text = element_text(size = 20)
  )

print(p)
```
