---
title: "Create Seurat object for Dec2024_pilot dataset"
editor_options: 
  chunk_output_type: console
---

# Import data

```{r setup}
SAVE_PLOTS = FALSE
SAVE_NEW_SEURAT_OBJECT <- TRUE
```

```{r load libs and functions}
#| message: false

library(Seurat)
library(tidyverse)
library(glue)
library(reticulate)
library(scCustomize)
library(patchwork)
library(future)

plan("sequential")
options(future.globals.maxSize = 400000 * 1024^2) # Allow large variables (300 GB)

source("03-scripts/R/seq_functions.R")
```

```{r create basic Seurat obj}
nuclei.dec <- readRDS('04-analysis/Seurats/Dec2024/seurat_curated.Rds')
nuclei.may <- readRDS('04-analysis/Seurats/May2024/seurat.Rds')
nuclei.may <- subset(nuclei.may, sublibrary == 1)

nuclei.dec$dataset <- 'Dec2024'
nuclei.may$dataset <- 'May2024'

# align barcode varname
nuclei.may$barcode  <- nuclei.may$bc_wells

nuclei <- merge(nuclei.dec, nuclei.may,
                add.cell.ids = c('dec', 'may'),
                merge.data = FALSE)

# nuclei[['RNA']] <- JoinLayers(nuclei[['RNA']])

# ensure no duplicate barcodes
nuclei$barcode <- paste(nuclei$barcode, nuclei$dataset, sep = '_')
```

Each sample has been SCT transformed independently. Try dimensionality reduction on this dataset to see how well they are together without integration
```{r add exp metadata}
nuclei <- nuclei |> 
  SCTransform(vars.to.regress = c('percent.mt', 'sublibrary'), 
              return.only.var.genes = TRUE,
              method = 'glmGamPoi',
              conserve.memory = TRUE)

nuclei <- nuclei |>
  RunPCA(npcs = 50, assay = 'SCT', seed.use = 17) |>
  FindNeighbors(dims = 1:41) |>
  FindClusters(resolution = 0.8, random.seed = 17) |>
  RunUMAP(dims = 1:41, seed.use = 17)

Idents(nuclei) <- nuclei$subclass_name
DimPlot(nuclei, group.by = 'dataset')
DimPlot(nuclei) +
  theme(legend.position = 'none') +
  scale_color_manual(values = LoadAllenColors('subclass'))
DimPlot(nuclei) +
  theme(legend.position = 'none') +
  scale_color_manual(values = LoadAllenColors('subclass'))
```


So the UMAP is showing 
# Intermediate processing

```{r visualize variable genes}
# Identify the 10 most highly variable genes
top50 <- head(VariableFeatures(nuclei_intermediate), 50)
p1 <- VariableFeaturePlot(nuclei_intermediate)
LabelPoints(p1, points = top50, repel = TRUE)
HoverLocator(p1)
```

```{r visualize PCs}
class_colors <- LoadAllenColors('class')
subclass_colors <- LoadAllenColors('subclass')
  
VizDimLoadings(nuclei_intermediate, dims = 1:5, reduction = "pca")

DimHeatmap(nuclei_intermediate, dims = 1:6, cells = 500, balanced = TRUE, fast = FALSE)
ElbowPlot(nuclei_intermediate,ndims = 50)
```

```{r visualize UMAP}
activity_colors <- LoadActivityColors("Dec2024")
subclass_colors <- LoadAllenColors()
ZT_colors <- LoadZTColors(5)

DimPlot(nuclei_intermediate, reduction = "pca", group.by = "class_name", raster = FALSE) + 
  scale_color_manual(values = class_colors) +
  theme(legend.position = 'none')

DimPlot(nuclei_intermediate, reduction = "pca", group.by = "subclass_name", raster = FALSE) + 
  scale_color_manual(values = subclass_colors) + 
  theme(legend.position = 'none')

DimPlot(nuclei_intermediate, reduction = "umap", group.by = "subclass_name", shuffle = TRUE, seed = 17, raster = FALSE) +
  scale_color_manual(values = subclass_colors) + 
  theme(legend.position = 'none')

DimPlot(nuclei_intermediate, reduction = "umap", group.by = "activity_condition", shuffle = TRUE, seed = 17, raster = FALSE) +
  scale_color_manual(values = activity_colors)
```

```{r save intermediate Seurat}
# SaveSeuratRds(nuclei_intermediate, file = paste0(save_dir, "seurat_intermediate.Rds")) # local
```

# Filter low-quality clusters

```{r static UMAPs}
# static UMAP feature plots
p1 <- FeaturePlot(nuclei_intermediate, reduction = 'umap', features = 'nCount_RNA', pt.size = 1, raster = F)
p2 <- FeaturePlot(nuclei_intermediate, reduction = 'umap', features = 'percent.mt', pt.size = 1, raster = F)

p1 / p2
```

```{r violin plots}
pUMI <- ggplot(nuclei_intermediate@meta.data) +
  aes(x = seurat_clusters, y = nCount_RNA, fill = seurat_clusters) +
  geom_violin(width = 2) +
  geom_boxplot(outliers = F, width = 0.2) +
  ylim(0, 30000) + 
  labs(y = 'UMI') +
  theme(legend.position = 'none',
        axis.title.y = element_text(size = 15),
        axis.title.x = element_blank())

pMt <- ggplot(nuclei_intermediate@meta.data) +
  aes(x = seurat_clusters, y = percent.mt, fill = seurat_clusters) +
  geom_jitter(shape = 21, width = 0.2, alpha = 0.2) +
  geom_boxplot(position = 'dodge', outliers = F, alpha = 0.8, width = 0.2) +
  labs(y = 'Percent mt reads') +
  theme(legend.position = 'none',
        axis.title.y = element_text(size = 15),
        axis.title.x = element_blank())

pD <- ggplot(nuclei_intermediate@meta.data) +
  aes(x = seurat_clusters, y = doublet_scores, fill = seurat_clusters) +
  geom_violin(width = 2) +
  geom_boxplot(outliers = F, width = 0.2) +
  labs(y = 'Doublet scores') +
  theme(legend.position = 'none',
        axis.title.y = element_text(size = 15),
        axis.title.x = element_blank())

pN <- nuclei_intermediate@meta.data |> 
  group_by(seurat_clusters) |> 
  summarize(n = n()) |> 
ggplot() +
  aes(x = seurat_clusters, y = n, fill = seurat_clusters) +
  geom_col() +
  scale_x_discrete(expand = expansion(mult = 0.013)) +
  scale_y_log10() +
  labs(y = 'log10(n)') +
  theme(legend.position = 'none',
        panel.grid.major.y = element_line(colour = "#808080"),
        axis.title.y = element_text(size = 15),
        axis.title.x = element_blank())

pUMI / pMt / pD / pN + 
  plot_layout(heights = c(7, 7, 7, 1)) +
  plot_annotation(title = 'Naive Seurat clustering of Dec2024')

# interactive plot for cross-referencing
DimPlot(nuclei_intermediate, 
        reduction = "umap", 
        group.by = "activity_condition", 
        shuffle = TRUE,
        seed = 17, 
        raster = FALSE) +
  scale_color_manual(values = activity_colors)

p1
p2

HoverLocator(
  DimPlot(nuclei_intermediate, 
        reduction = "umap", 
        group.by = "subclass_name", 
        shuffle = TRUE,
        seed = 17, 
        raster = FALSE) +
  scale_color_manual(values = subclass_colors),
  information = nuclei_intermediate@meta.data
)
```

The above graph has the following takeaways:

-   Based on UMI count, the clusters that could be discarded:
    -   1, legit astrocytes -- keep
    -   2, legit oligos -- keep
    -   4, kegit OPCs -- keep
    -   15, legit OPC cluster -- keep
    -   23, vascular cells -- keep
    -   26, IMN, cajal-retzius, and astrocytes continuum -- keep
    -   29, 6h KA astrocytes -- keep
    -   37, real astrocytes -- keep
    -   40, more cajal-retzius -- keep
    -   47, random astrocytes -- delete

-   Based on mt read count, the clusters that could be discarded:
    -   15, legit OPC cluster -- keep
    -   21, legit CA1 6h KA cluster -- keep
    -   23, vascular cells -- keep
    -   29, 6h KA astrocytes -- keep
    -   35, all over the place -- delete
    -   37, real astrocytes -- keep
    -   43, random astrocytes and oligos -- delete
    -   47, random astrocytes -- delete

-   Based on the doublet scores, the clusters that could be discarded:
    -   20, CA1 cluster -- keep
    -   21, legit CA1 6h KA cluster -- keep
    -   24, somewhat large cluster of subiculum and CA1 -- delete
    -   36, all over the place -- delete
    -   43, random astrocytes and oligos -- delete
    -   44, unsure of class right now -- keep
    -   45, random CA1 -- delete
    -   47, random astrocytes -- delete

-   So the compiled list of clusters to delete is:
    -   24
    -   35
    -   36
    -   43
    -   45
    -   47

```{r isolate clusters in UMAP}
clusters_to_plot <- c(
  1,
  2,
  4,
  26,
  40
  )

for (cluster in clusters_to_plot) {
  p <- DimPlot(nuclei_intermediate, reduction = 'umap', group.by = 'seurat_clusters', pt.size = 1, raster = FALSE) + 
    scale_color_manual(values = setNames('red', cluster)) +
    labs(title = glue("cluster {cluster}")) +
    theme(plot.title = element_text(size = 30),
          legend.position = 'none')

  print(p)
}
```

## Remove low-quality clusters

```{r}
clusters_to_remove <- c(
  24,
  35,
  36,
  43,
  45,
  47
)

# get list of accepted barcodes NOT in bad clusters
bc_to_keep <- nuclei_intermediate@meta.data |> 
  filter(!seurat_clusters %in% clusters_to_remove) |> 
  pull(barcode)

# remove bad clusters and doublets
nuclei_curated <- nuclei_intermediate |> 
  subset(barcode %in% bc_to_keep) |> 
  subset(doublet_scores <= 0.02)

# save this part of the process
SaveSeuratRds(object = nuclei_curated, file = paste0(save_dir, "seurat_curated.Rds"))
```

## Re-do intermediate processing object

```{r process dataset with Seurat}
nuclei <- nuclei_curated |> 
  RunPCA(npcs = 50) |> 
  FindNeighbors(dims = 1:41) |> 
  FindClusters(resolution = 0.8, random.seed = 17) |> 
  RunUMAP(dims = 1:41, seed.use = 17)

# make cell identities default to subclass_name
Idents(nuclei) <- nuclei$subclass_name
```

```{r visualize PCs}
DimHeatmap(nuclei, dims = 1:6, cells = 500, balanced = TRUE, fast = FALSE)
ElbowPlot(nuclei ,ndims = 50)
```

```{r visualize UMAP}
DimPlot(nuclei, reduction = "umap", group.by = "activity_condition", shuffle = TRUE, seed = 17, pt.size = 1, raster = F) +
  scale_color_manual(values = activity_colors) +
  theme(legend.position = 'none')

p1 <- DimPlot(nuclei, reduction = "umap", group.by = "subclass_name", shuffle = TRUE, seed = 17, pt.size = 1, raster = F) +
  scale_color_manual(values = subclass_colors) +
  theme(legend.position = 'none')

HoverLocator(p1, information = nuclei_curated@meta.data)
```

# Re-join layers

```{r}
nuclei <- JoinLayers(nuclei)
```

# Save object

```{r save Seurat object}
if (SAVE_NEW_SEURAT_OBJECT) {
  print('Saving Seurat object...')
  
  SaveSeuratRds(nuclei, file = paste0(save_dir, "seurat.Rds")) # local
  
  print('Done!')
}
```
