---
title: "Create Seurat object for Dec2024_pilot dataset"
editor_options: 
  chunk_output_type: console
---

# Import data

```{r setup}
SAVE_PLOTS = FALSE
SAVE_NEW_SEURAT_OBJECT <- TRUE
```

```{r load libs and functions}
#| message: false

library(Seurat)
library(tidyverse)
library(glue)
library(reticulate)
library(scCustomize)
library(patchwork)
library(future)

plan("sequential")
options(future.globals.maxSize = 400000 * 1024^2) # Allow large variables (300 GB)

source("03-scripts/R/seq_functions.R")
```

```{r create basic Seurat obj}
dataset <- 'Dec2024'

save_dir <- glue("04-analysis/Seurats/{dataset}/")

# make save_dir if it doesn't exist
if (!dir.exists(save_dir)) {
  dir.create(save_dir, recursive = TRUE)
}

nuclei_trailmaker_unfiltered <- readRDS("04-analysis/Seurats/Dec2024/seurat_trailmaker_Dec2024_complete.rds")
```

# Rename Features

Very annoying fact: Seurat objects downloaded from Trailmaker use unrecognizable Ensembl IDs for feature names rather than gene symbols. In order to replace the Feature names, I have to take a circuitous route by creating an entirely new Seurat object. The 2 components of the new Seurat object are the counts matrix and the metadata. The features for the object will come from the row names of the counts matrix. To set the row names, I have to map gene symbols from Ensembl IDs using a key provided through Trailmaker. For rows where the gene symbol maps to multiple Ensembl IDs, I only keep the first row with that gene symbol and discard the other rows.

```{r rename Features}
gene_dict <- read_csv("02-data/raw_data/all_genes.csv")
gene_dict <- setNames(gene_dict$gene_name, gene_dict$gene_id)

old_genes <- rownames(nuclei_trailmaker_unfiltered)
new_genes <- gene_dict[old_genes]  # map old to new names

new_genes[is.na(new_genes)]  # check new names

# Extract raw count matrix
counts_mat <- nuclei_trailmaker_unfiltered@assays$RNA@counts

# Get the first occurrence of each unique gene symbol
first_occurrence_idx <- match(unique(new_genes), new_genes)

# Subset the matrix to retain only the first occurrence
filtered_counts <- counts_mat[first_occurrence_idx, ]

# Assign row names as unique gene symbols
rownames(filtered_counts) <- unique(new_genes)
colnames(filtered_counts) <- colnames(nuclei_trailmaker_unfiltered)

nuclei_trailmaker_unfiltered <- CreateSeuratObject(counts = filtered_counts, meta.data = nuclei_trailmaker_unfiltered@meta.data)

rm(counts_mat, filtered_counts)
```

```{r add exp metadata}
cell_meta <- nuclei_trailmaker_unfiltered@meta.data |> 
  select(-ends_with(c('_well', '_wind'))) |> 
  select(-starts_with('emptyDrops')) |> 
  select(-starts_with('custom_cellset')) |> 
  rename(sample = samples) |> 
  mutate(sample = str_replace(sample, '30M', '30m')) |> 
  mutate(sample = str_replace(sample, '6H', '6h')) |> 
  relocate(sample, .after = barcode)

experimental_meta_path <- "01-documentation/sample_experimental_metadata_Dec2024.csv"
experimental_meta <- read_csv(experimental_meta_path)

# merge with metadata by sample
meta_unfiltered <- merge(cell_meta, experimental_meta) |> 
  # LINE IS CRITICAL TO ASSIGN CORRECT METADATA TO CELLS-- DO NOT DELETE
  arrange(match(barcode, rownames(nuclei_trailmaker_unfiltered@meta.data)))

# add metadata to seurat
nuclei_trailmaker_unfiltered <- AddMetaData(nuclei_trailmaker_unfiltered, meta_unfiltered)
```

# Cutoff filtering

```{r QC filter cells}
nuclei_trailmaker_unfiltered$percent.mt <- PercentageFeatureSet(nuclei_trailmaker_unfiltered, pattern = "^mt-")

nuclei_trailmaker_filtered <- nuclei_trailmaker_unfiltered |> 
  subset(
    nCount_RNA >= 1500 &
    nFeature_RNA >= 1500 & 
    percent.mt <= 5
  )
```

```{r}
as.anndata(x = nuclei_trailmaker_filtered, file_path = save_dir, file_name = "nuclei_MapMyCells", main_layer = "counts", other_layers = NULL)
```

```{r merge metadata sources}
# load in Allen metadata
allen_meta_path = "04-analysis/Seurats/Dec2024/nuclei_MapMyCells_10xWholeMouseBrain(CCN20230722)_HierarchicalMapping_UTC_1738089323143/nuclei_MapMyCells_10xWholeMouseBrain(CCN20230722)_HierarchicalMapping_UTC_1738089323143.csv"
allen_meta <- read_csv(allen_meta_path, skip = 4)

# meta_merged <- merge(nuclei_filtered@meta.data, allen_meta, by.x = 'bc_wells', by.y = 'cell_id')
meta_merged <- left_join(nuclei_trailmaker_filtered@meta.data, allen_meta, by = join_by('barcode' == 'cell_id'))
```

# Add metadata

```{r mutate meta variables}
# make new columns
meta <- meta_merged |>
  mutate(sublibrary = factor(str_sub(barcode, -1))) |>
  mutate(ZT = str_extract(sample, "^[^_]+")) |>
  mutate(seizure = ifelse(str_detect(ZT, "KA"), TRUE, FALSE)) |>
  mutate(ZT = ifelse(seizure == T, NA, ZT))

# make col: activity_condition (e.g. SE, EE6h, KA30m)
meta <- meta |>
  mutate(replicate = str_sub(sample, -1)) |> 
  mutate(activity_condition = str_sub(sample, 1, -2)) |>  # remove replicate numbers
  mutate(activity_condition = ifelse(
    str_detect(str_sub(activity_condition, 1, 2), "KA"),         # if KA
    str_remove(activity_condition, '_'),                         # then make it KA30m or KA6h
    str_remove(str_extract(activity_condition, "(?<=_).*"), '_') # else only include text after first underscore
  )) |> 
  mutate(activity_condition = ifelse(         # add "EE_" prefix to 30m and 6h
    str_detect(activity_condition, "SE"), 
    "SE", 
    ifelse(
      !str_detect(activity_condition, "KA"), 
      paste0("EE", activity_condition), 
      activity_condition
  )))
  

meta$ZT <- factor(meta$ZT, levels = c('ZT0', 'ZT4', 'ZT12', 'ZT16'))
meta$sublibrary <- factor(meta$sublibrary, levels = c('1', '2', '3', '4', '5', '6', '7', '8'))
meta$activity_condition <- factor(meta$activity_condition, levels = c('SE', 'EE30m', 'EE6h', 'KA30m', 'KA6h'))
meta$sample <- factor(meta$sample, levels = c(
  'KA_30m1',
  'KA_30m2',
  'KA_30m3',
  'KA_6h1',
  'KA_6h2',
  'KA_6h3',
  
  'ZT0_SE1',
  'ZT0_SE2',
  'ZT0_SE3',
  'ZT0_SE4',
  'ZT0_30m1',
  'ZT0_30m2',
  'ZT0_30m3',
  'ZT0_30m4',
  'ZT0_6h1',
  'ZT0_6h2',
  'ZT0_6h3',
  
  'ZT4_SE1',
  'ZT4_SE2',
  'ZT4_SE3',
  'ZT4_SE4',
  'ZT4_30m1',
  'ZT4_30m2',
  'ZT4_30m3',
  'ZT4_30m4',
  'ZT4_6h1',
  'ZT4_6h2',
  'ZT4_6h3',
  
  'ZT12_SE1',
  'ZT12_SE2',
  'ZT12_SE3',
  'ZT12_30m1',
  'ZT12_30m2',
  'ZT12_30m3',
  'ZT12_6h1',
  'ZT12_6h2',
  'ZT12_6h3',
  
  'ZT16_SE1',
  'ZT16_SE2',
  'ZT16_SE3',
  'ZT16_SE4',
  'ZT16_30m1',
  'ZT16_30m2',
  'ZT16_30m3',
  'ZT16_30m4',
  'ZT16_6h1',
  'ZT16_6h2',
  'ZT16_6h3'
))

# add new info, re-order cols, remove useless cols
meta <- meta |> 
  relocate(sublibrary, sample, ZT, activity_condition, .after = barcode) |>
  relocate(ends_with("_name"), .after = percent.mt) |>
  # THIS LINE IS VERY IMPORTANT SO CELLS GET RIGHT METADATA
  arrange(match(barcode, rownames(nuclei_trailmaker_filtered@meta.data)))

nuclei_intermediate <- AddMetaData(nuclei_trailmaker_filtered, meta)
```

# Intermediate processing

Normalize data, build UMAP embedding.

```{r process dataset with Seurat}
# split object
# nuclei_intermediate[['RNA']] <- split(nuclei_intermediate[['RNA']], f = nuclei_intermediate$sample)

nuclei_intermediate <- nuclei_intermediate |> 
  # NormalizeData(normalization.method = 'LogNormalize', scale.factor = 10000) |> 
  # FindVariableFeatures() |> 
  # ScaleData() |> 
  SCTransform(vars.to.regress = c('percent.mt', 'sublibrary'), 
              return.only.var.genes = FALSE,
              method = 'glmGamPoi',
              conserve.memory = FALSE)

# save nuclei_intermediate
SaveSeuratRds(nuclei_intermediate, file = paste0(save_dir, "seurat_intermediate.Rds")) # local

# keep going
nuclei_intermediate <- nuclei_intermediate |> 
  RunPCA(npcs = 50) |> 
  FindNeighbors(dims = 1:41) |> 
  FindClusters(resolution = 0.8, random.seed = 17) |> 
  RunUMAP(dims = 1:41, seed.use = 17)

# make cell identities default to subclass_name
Idents(nuclei_intermediate) <- nuclei_intermediate$subclass_name
rownames(nuclei_intermediate@meta.data) <- Cells(nuclei_intermediate)
```

```{r visualize variable genes}
# Identify the 10 most highly variable genes
top50 <- head(VariableFeatures(nuclei_intermediate), 50)
p1 <- VariableFeaturePlot(nuclei_intermediate)
LabelPoints(p1, points = top50, repel = TRUE)
HoverLocator(p1)
```

```{r visualize PCs}
class_colors <- LoadAllenColors('class')
subclass_colors <- LoadAllenColors('subclass')
  
VizDimLoadings(nuclei_intermediate, dims = 1:5, reduction = "pca")

DimHeatmap(nuclei_intermediate, dims = 1:6, cells = 500, balanced = TRUE, fast = FALSE)
ElbowPlot(nuclei_intermediate,ndims = 50)
```

```{r visualize UMAP}
activity_colors <- LoadActivityColors("Dec2024")
subclass_colors <- LoadAllenColors()
ZT_colors <- LoadZTColors(5)

DimPlot(nuclei_intermediate, reduction = "pca", group.by = "class_name", raster = FALSE) + 
  scale_color_manual(values = class_colors) +
  theme(legend.position = 'none')

DimPlot(nuclei_intermediate, reduction = "pca", group.by = "subclass_name", raster = FALSE) + 
  scale_color_manual(values = subclass_colors) + 
  theme(legend.position = 'none')

DimPlot(nuclei_intermediate, reduction = "umap", group.by = "subclass_name", shuffle = TRUE, seed = 17, raster = FALSE) +
  scale_color_manual(values = subclass_colors) + 
  theme(legend.position = 'none')

DimPlot(nuclei_intermediate, reduction = "umap", group.by = "activity_condition", shuffle = TRUE, seed = 17, raster = FALSE) +
  scale_color_manual(values = activity_colors)
```

```{r save intermediate Seurat}
# SaveSeuratRds(nuclei_intermediate, file = paste0(save_dir, "seurat_intermediate.Rds")) # local
```

# Filter low-quality clusters

```{r static UMAPs}
# static UMAP feature plots
p1 <- FeaturePlot(nuclei_intermediate, reduction = 'umap', features = 'nCount_RNA', pt.size = 1, raster = F)
p2 <- FeaturePlot(nuclei_intermediate, reduction = 'umap', features = 'percent.mt', pt.size = 1, raster = F)

p1 / p2
```

```{r violin plots}
pUMI <- ggplot(nuclei_intermediate@meta.data) +
  aes(x = seurat_clusters, y = nCount_RNA, fill = seurat_clusters) +
  geom_violin(width = 2) +
  geom_boxplot(outliers = F, width = 0.2) +
  ylim(0, 30000) + 
  labs(y = 'UMI') +
  theme(legend.position = 'none',
        axis.title.y = element_text(size = 15),
        axis.title.x = element_blank())

pMt <- ggplot(nuclei_intermediate@meta.data) +
  aes(x = seurat_clusters, y = percent.mt, fill = seurat_clusters) +
  geom_jitter(shape = 21, width = 0.2, alpha = 0.2) +
  geom_boxplot(position = 'dodge', outliers = F, alpha = 0.8, width = 0.2) +
  labs(y = 'Percent mt reads') +
  theme(legend.position = 'none',
        axis.title.y = element_text(size = 15),
        axis.title.x = element_blank())

pD <- ggplot(nuclei_intermediate@meta.data) +
  aes(x = seurat_clusters, y = doublet_scores, fill = seurat_clusters) +
  geom_violin(width = 2) +
  geom_boxplot(outliers = F, width = 0.2) +
  labs(y = 'Doublet scores') +
  theme(legend.position = 'none',
        axis.title.y = element_text(size = 15),
        axis.title.x = element_blank())

pN <- nuclei_intermediate@meta.data |> 
  group_by(seurat_clusters) |> 
  summarize(n = n()) |> 
ggplot() +
  aes(x = seurat_clusters, y = n, fill = seurat_clusters) +
  geom_col() +
  scale_x_discrete(expand = expansion(mult = 0.013)) +
  scale_y_log10() +
  labs(y = 'log10(n)') +
  theme(legend.position = 'none',
        panel.grid.major.y = element_line(colour = "#808080"),
        axis.title.y = element_text(size = 15),
        axis.title.x = element_blank())

pUMI / pMt / pD / pN + 
  plot_layout(heights = c(7, 7, 7, 1)) +
  plot_annotation(title = 'Naive Seurat clustering of Dec2024')

# interactive plot for cross-referencing
DimPlot(nuclei_intermediate, 
        reduction = "umap", 
        group.by = "activity_condition", 
        shuffle = TRUE,
        seed = 17, 
        raster = FALSE) +
  scale_color_manual(values = activity_colors)

p1
p2

HoverLocator(
  DimPlot(nuclei_intermediate, 
        reduction = "umap", 
        group.by = "subclass_name", 
        shuffle = TRUE,
        seed = 17, 
        raster = FALSE) +
  scale_color_manual(values = subclass_colors),
  information = nuclei_intermediate@meta.data
)
```

The above graph has the following takeaways:

-   Based on UMI count, the clusters that could be discarded:
    -   1, legit astrocytes -- keep
    -   2, legit oligos -- keep
    -   4, kegit OPCs -- keep
    -   15, legit OPC cluster -- keep
    -   23, vascular cells -- keep
    -   26, IMN, cajal-retzius, and astrocytes continuum -- keep
    -   29, 6h KA astrocytes -- keep
    -   37, real astrocytes -- keep
    -   40, more cajal-retzius -- keep
    -   47, random astrocytes -- delete
-   Based on mt read count, the clusters that could be discarded:
    -   15, legit OPC cluster -- keep
    -   21, legit CA1 6h KA cluster -- keep
    -   23, vascular cells -- keep
    -   29, 6h KA astrocytes -- keep
    -   35, all over the place -- delete
    -   37, real astrocytes -- keep
    -   43, random astrocytes and oligos -- delete
    -   47, random astrocytes -- delete
-   Based on the doublet scores, the clusters that could be discarded:
    -   20, CA1 cluster -- keep
    -   21, legit CA1 6h KA cluster -- keep
    -   24, somewhat large cluster of subiculum and CA1 -- delete
    -   36, all over the place -- delete
    -   43, random astrocytes and oligos -- delete
    -   44, unsure of class right now -- keep
    -   45, random CA1 -- delete
    -   47, random astrocytes -- delete
-   So the compiled list of clusters to delete is:
    -   24
    -   35
    -   36
    -   43
    -   45
    -   47

```{r isolate clusters in UMAP}
clusters_to_plot <- c(
  1,
  2,
  4,
  26,
  40
  )

for (cluster in clusters_to_plot) {
  p <- DimPlot(nuclei_intermediate, reduction = 'umap', group.by = 'seurat_clusters', pt.size = 1, raster = FALSE) + 
    scale_color_manual(values = setNames('red', cluster)) +
    labs(title = glue("cluster {cluster}")) +
    theme(plot.title = element_text(size = 30),
          legend.position = 'none')

  print(p)
}
```

## Remove low-quality clusters

```{r}
clusters_to_remove <- c(
  24,
  35,
  36,
  43,
  45,
  47
)

# get list of accepted barcodes NOT in bad clusters
bc_to_keep <- nuclei_intermediate@meta.data |> 
  filter(!seurat_clusters %in% clusters_to_remove) |> 
  pull(barcode)

# remove bad clusters and doublets
nuclei_curated <- nuclei_intermediate |> 
  subset(barcode %in% bc_to_keep) |> 
  subset(doublet_scores <= 0.02)

# save this part of the process
SaveSeuratRds(object = nuclei_curated, file = paste0(save_dir, "seurat_curated.Rds"))
```

## Re-do intermediate processing object

```{r process dataset with Seurat again}
nuclei <- nuclei_curated |> 
  RunPCA(npcs = 50) |> 
  FindNeighbors(dims = 1:41) |> 
  FindClusters(resolution = 0.8, random.seed = 17) |> 
  RunUMAP(dims = 1:41, seed.use = 17)

# make cell identities default to subclass_name
Idents(nuclei) <- nuclei$subclass_name
rownames(nuclei@meta.data) <- Cells(nuclei)
```

```{r visualize PCs again}
DimHeatmap(nuclei, dims = 1:6, cells = 500, balanced = TRUE, fast = FALSE)
ElbowPlot(nuclei ,ndims = 50)
```

```{r visualize UMAP again}
DimPlot(nuclei, reduction = "umap", group.by = "activity_condition", shuffle = TRUE, seed = 17, pt.size = 1, raster = F) +
  scale_color_manual(values = activity_colors) +
  theme(legend.position = 'none')

p1 <- DimPlot(nuclei, reduction = "umap", group.by = "subclass_name", shuffle = TRUE, seed = 17, pt.size = 1, raster = F) +
  scale_color_manual(values = subclass_colors) +
  theme(legend.position = 'none')

HoverLocator(p1, information = nuclei_curated@meta.data)
```

# Re-join layers

```{r}
nuclei <- JoinLayers(nuclei)
```

# Save object

```{r save Seurat object}
if (SAVE_NEW_SEURAT_OBJECT) {
  print('Saving Seurat object...')
  
  SaveSeuratRds(nuclei, file = paste0(save_dir, "seurat.Rds")) # local
  
  print('Done!')
}
```
