---
title: "Manuscript DGE analysis"
editor_options: 
  chunk_output_type: inline
---

This code iterates through all major subclasses represented in the data and finds the differentially expressed genes in every contrast combination. It uses MAST and has latent variables "sublibrary" and "percent.mt" as variables. Finally, any sex-specific genes (i.e., those differentially expressed in the Yao et al. 2023 dataset) that show up in the DEG list are removed.

**Differential gene expression methods:**

Differentially expressed genes between different conditions were determined using the MAST algorithm called inside Seurat's `FindMarkers()` function, with the following parameters included: `FindMarkers(logfc.threshold = 0.25, test.use = 'MAST', latent.vars = c('sublibrary', 'percent.mt'))`. Both nucleus sublibrary and the percent of reads mapping to mitochondrial genes were included as latent variables in the MAST model. Sexually dimorphically expressed genes were found by leveraging the large dataset published in Yao et al. 2023. Instead of including conditions as groups in the FindMarkers() function, sex of the donor animal was used. Both sexually dimorphic genes and between-condition DEGs were calculated for each subclass independently. For each subclass, any gene symbols discovered in both the between-condition and between-sex DEG lists were discarded from the between-condition gene list in order to remove genes that were differentially expressed on the basis of animal sex.

```{r load libs & data}
library(Seurat)
library(SeuratDisk)

library(tidyverse)

library(reshape2)

library(hues)
library(paletteer)
library(patchwork)

source("03-scripts/R/seq_functions.R")

nuclei <- LoadDataset("Dec2024")

csv_write_dir <- "04-analysis/DEGs/Dec2024_pilot_activity_condition"
```

```{r select major subclasses}
subclass_list <- LoadSubclassesToUse(nuclei) |> 
  print() |>  # print n before pulling just the list of names
  pull(subclass_name)

nuclei <- nuclei |> 
  subset(subclass_name %in% subclass_list)
```

```{r removeSexSpecificGenes()}
# removes any sex-specific genes found in the Yao dataset from DEG list
removeSexSpecificGenes <- function(subclass, naive_gene_list) {
  print(glue("Removing sex-specific genes from {subclass}"))
  
  # prep subclass names
  subclass_fname <- gsub(" ", "_", subclass)
  subclass_fname <- gsub("/", "", subclass_fname)
  
  # load in sex-specific genes for this subclass
  fname <- glue("04-analysis/DEGs/Yao2023_sex_DEGs/{subclass_fname}_DEGs.csv")
  sex_gene_list <- read_csv(fname) |> 
    pull(gene)
  
  # remove any genes from naive_gene_list found in sex_gene_list
  scrubbed_gene_list <- naive_gene_list[!naive_gene_list %in% sex_gene_list]
  
  return(scrubbed_gene_list)
}
```

```{r findDEG()}
# # uses MAST to find DEGs between conditions within a subclass and removes sex-specific DEGs
# findDEG <- function(subclass, group1, group2, logFC_threshold=0.25) {
#   # print report statement
#   print(glue("{subclass} between {group1} and {group2}."))
  
#   # subset nuclei
#   nuclei_subclass <- subset(nuclei, subclass_name == subclass)
#   Idents(nuclei_subclass) <- nuclei_subclass$condition
  
#   # find DEGs
#   DEGs <- FindMarkers(
#     nuclei_subclass, 
#     logfc.threshold = logFC_threshold,
#     test.use = 'MAST',
#     ident.1 = group1, ident.2 = group2,
#     latent.vars = c('sublibrary', 'percent.mt'),
#   ) |> 
#     filter(p_val_adj < 0.05) |> 
#     rownames_to_column(var = 'gene')
  
#   # remove sex-specific genes
#   naive_DEG_list <- DEGs |> 
#     pull(gene)
#   # clean_DEG_list <- removeSexSpecificGenes(subclass, naive_DEG_list)
#   clean_DEG_list <- naive_DEG_list
#   DEGs <- DEGs |> 
#     filter(gene %in% clean_DEG_list)
  
#   return(DEGs)
# }
```

```{r find DEGs for subclasses, message = FALSE}
log2FC_threshold <- 0.585  # corresponding to 50% change

for (i in seq_along(subclass_list)) {
  subclass <- subclass_list[i]
  print(glue("subclass {i}/{length(subclass_list)}"))
  
  subclass_fname <- ShrinkSubclassName(subclass)

  DGE_EE30m_vs_SE <- FindDEGs(nuclei, subclass, 'activity_condition', 'EE_30m', 'SE', log2FC_threshold)
  write_csv(DGE_EE30m_vs_SE, glue("{csv_write_dir}/{subclass_fname}_EE30m_vs_SE.csv"))
  
  DGE_EE6h_vs_SE <- FindDEGs(nuclei, subclass, 'activity_condition', 'EE_6h', 'SE', log2FC_threshold)
  write_csv(DGE_EE6h_vs_SE, glue("{csv_write_dir}/{subclass_fname}_EE6h_vs_SE.csv"))
  
  DGE_KA30m_vs_SE <- FindDEGs(nuclei, subclass, 'activity_condition', 'KA_30m', 'SE', log2FC_threshold)
  write_csv(DGE_KA30m_vs_SE, glue("{csv_write_dir}/{subclass_fname}_KA30m_vs_SE.csv"))
  
  DGE_KA6h_vs_SE <- FindDEGs(nuclei, subclass, 'activity_condition', 'KA_6h', 'SE', log2FC_threshold)
  write_csv(DGE_KA6h_vs_SE, glue("{csv_write_dir}/{subclass_fname}_KA6h_vs_SE.csv"))
  
  DGE_KA30m_vs_EE30m <- FindDEGs(nuclei, subclass, 'activity_condition', 'KA_30m', 'EE_30m', log2FC_threshold)
  write_csv(DGE_KA30m_vs_EE30m, glue("{csv_write_dir}/{subclass_fname}_KA30m_vs_EE30m.csv"))
  
  DGE_KA6h_vs_EE6h <- FindDEGs(nuclei, subclass, 'activity_condition', 'KA_6h', 'EE_6h', log2FC_threshold)
  write_csv(DGE_KA6h_vs_EE6h, glue("{csv_write_dir}/{subclass_fname}_KA6h_vs_EE6h.csv"))
  
}
```

```{r}
# this contrast was errantly calcualted in the previous loop 
for (i in seq_along(subclass_list)) {
  subclass <- subclass_list[i]
  print(glue("subclass {i}/{length(subclass_list)}"))
  
  DGE_KA6h_vs_EE6h <- FindDEGs(nuclei, subclass, 'activity_condition', 'KA_6h', 'EE_6h', log2FC_threshold)
  write_csv(DGE_KA6h_vs_EE6h, glue("{csv_write_dir}/{subclass_fname}_KA6h_vs_EE6h.csv"))
}
```
