---
title: "doublet pilot analysis"
editor_options: 
  chunk_output_type: console
---

The general plan for this script is to pilot doublet detection on my May2024 dataset. I will use the DoubletFinder package to find doublets in my **unfiltered** May2024 dataset, and filter out those cells. I will then compare the cells filtered out with DoubletFinder to the cells filtered out using normal thresholding.

```{r}
library(Seurat)
library(DoubletFinder)
library(tidyverse)
library(patchwork)
library(scCustomize)
library(venndir)

library(reticulate)
use_condaenv('actdep')

source("03-scripts/R/seq_functions.R")
```

```{r}
nuclei_filtered <- LoadDataset("May2024")
bc_filtered <- rownames(nuclei_filtered@meta.data) |> str_sub(1, 8)
```

```{r}
data_dir <- "/Volumes/jack/seq/analysis_May2024/sublibrary1/0_all-sample/DGE_unfiltered/"
mat_unfiltered <- ReadParseBio(data_dir)
meta_unfiltered <- read_csv(paste0(data_dir, 'cell_metadata.csv'))
nuclei_unfiltered_raw <- CreateSeuratObject(counts = mat_unfiltered, meta.data = meta_unfiltered)
```

```{r threshold-filter}
nuclei_unfiltered_raw$percent.mt <- PercentageFeatureSet(nuclei_unfiltered_raw, pattern = "^mt-")

# only use lower thresholds here
nuclei_unfiltered_raw <- nuclei_unfiltered_raw |> 
  subset(
    tscp_count >= 1500 &
    gene_count >= 1500 &
    percent.mt <= 1
  )

bc_unfiltered <- rownames(nuclei_unfiltered_raw@meta.data)
```

```{r process Seurat}
options(future.globals.maxSize = 14000 * 1024^2) # allow variables to be 14 GB large

nuclei_unfiltered <- nuclei_unfiltered_raw |> 
  NormalizeData(normalization.method = 'LogNormalize', scale.factor = 10000) |> 
  FindVariableFeatures() |> 
  ScaleData() |> 
  RunPCA() |> 
  FindNeighbors(dims = 1:41) |> 
  FindClusters(resolution = 0.8) |>
  BuildClusterTree(reorder = TRUE, reorder.numeric = "TRUE") |>
  RunUMAP(dims = 1:41)
```

```{r}
# save the unfiltered seurat in-place so we can get Allen labels
as.anndata(x = nuclei_unfiltered, file_path = data_dir, file_name = 'Seurat_doublet_piloting', main_layer = 'counts', other_layers = NULL)
```

```{r merge Allen meta}
meta_allen <- read_csv("/Volumes/jack/seq/analysis_May2024/sublibrary1/0_all-sample/DGE_unfiltered/Seurat_doublet_piloting_10xWholeMouseBrain(CCN20230722)_HierarchicalMapping_UTC_1735580806722/Seurat_doublet_piloting_10xWholeMouseBrain(CCN20230722)_HierarchicalMapping_UTC_1735580806722.csv", skip = 4)

meta_raw <- nuclei_unfiltered@meta.data
meta <- meta_raw |> 
  left_join(meta_allen, by = c('bc_wells' = 'cell_id'))
rownames(meta) <- meta$bc_wells

nuclei_unfiltered <- AddMetaData(nuclei_unfiltered, meta)

# remove cells in small subclasses
subclasses_to_use <- LoadSubclassesToUse(seurat_obj = nuclei_unfiltered, cell_cutoff = 10)
cells_to_keep <- meta |> 
  filter(subclass_name %in% subclasses_to_use$subclass_name) |> 
  pull(bc_wells)
nuclei_unfiltered <- subset(nuclei_unfiltered, bc_wells %in% cells_to_keep)
```

Perform doublet detection on the unfiltered object

```{r pK identification}
sweep_results <- paramSweep(nuclei_unfiltered, PCs = 1:10, sct = FALSE)
sweep_stats <- summarizeSweep(sweep_results, GT = FALSE)
pK_table <- find.pK(sweep_stats)
```

```{r homotypic doublet rate}
homotypic.prop <- modelHomotypic(nuclei_filtered$seurat_clusters)
nExp_poi <- round(0.075*nrow(nuclei_unfiltered@meta.data))
nExp_poi.adj <- round(nExp_poi*(1-homotypic.prop))
```

```{r find doublets}
nuclei_unfiltered <- doubletFinder(nuclei_unfiltered,
              sct = FALSE,
              PCs = 1:30,
              pK = 0.01,
              nExp = nExp_poi
              )
nuclei_unfiltered$doublet.class <- factor(nuclei_unfiltered$DF.classifications_0.25_0.01_452, levels = c('Singlet', 'Doublet'))
bc_doublets <- names(nuclei_unfiltered$doublet.class[nuclei_unfiltered$doublet.class == 'Doublet'])
```

```{r plot doublets}
# doublet plot
p <- DimPlot(nuclei_unfiltered, 
            group.by = 'doublet.class', 
            pt.size = 1) + 
  scale_color_manual(values = setNames(c('black','red'), c('Singlet', 'Doublet')))
HoverLocator(p, information = nuclei_unfiltered@meta.data)

# subclass plot unfiltered
subclass_colors <- LoadAllenColors()
p <- DimPlot(nuclei_unfiltered,
            group.by = 'subclass_name',
            pt.size = 1) +
  scale_color_manual(values = subclass_colors) +
  theme(legend.position = 'none')
HoverLocator(p, information = nuclei_unfiltered@meta.data)

# subclass plot filtered
p <- DimPlot(nuclei_filtered,
            group.by = 'subclass_name',
            pt.size = 1) +
  scale_color_manual(values = subclass_colors) +
  theme(legend.position = 'none')
HoverLocator(p, information = nuclei_filtered@meta.data)
```

Now that we have labels for Singlets and Doublets, let's explore how these classes differ

```{r}
# read plot
ggplot(nuclei_unfiltered@meta.data) +
  aes(x = doublet.class, y = mread_count, fill = doublet.class) +
  geom_jitter(shape = 21, width = 0.2) +
  geom_boxplot(width = 0.25, alpha = 0.5, color = 'white', outliers = F) + 
  scale_fill_manual(values = setNames(c('black','red'), c('Singlet', 'Doublet'))) +
  facet_wrap(~subclass_name, scales = 'free') +
  labs(title = 'Read count: singlets vs doublets across clusters', y = 'Read count') +
  theme(
    plot.title = element_text(size = 25),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 15),
    axis.text.y = element_text(size = 12),
    axis.text.x = element_text(size = 15),
    legend.position = 'none'
  )

# transcript plot
ggplot(nuclei_unfiltered@meta.data) +
  aes(x = doublet.class, y = tscp_count, fill = doublet.class) +
  geom_jitter(shape = 21, width = 0.2) +
  geom_boxplot(width = 0.25, alpha = 0.5, color = 'white', outliers = F) + 
  scale_fill_manual(values = setNames(c('black','red'), c('Singlet', 'Doublet'))) +
  facet_wrap(~subclass_name, scales = 'free') +
  labs(title = 'Transcript count: singlets vs doublets across clusters', y = 'Transcript count') +
  theme(
    plot.title = element_text(size = 25),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 15),
    axis.text.y = element_text(size = 12),
    axis.text.x = element_text(size = 15),
    legend.position = 'none'
  )

# gene plot
ggplot(nuclei_unfiltered@meta.data) +
  aes(x = doublet.class, y = gene_count, fill = doublet.class) +
  geom_jitter(shape = 21, width = 0.2) +
  geom_boxplot(width = 0.25, alpha = 0.5, color = 'white', outliers = F) + 
  scale_fill_manual(values = setNames(c('black','red'), c('Singlet', 'Doublet'))) +
  facet_wrap(~subclass_name, scales = 'free') +
  labs(title = 'Gene count: singlets vs doublets across clusters', y = 'Gene count') +
  theme(
    plot.title = element_text(size = 25),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 15),
    axis.text.y = element_text(size = 12),
    axis.text.x = element_text(size = 15),
    legend.position = 'none'
  )
```

Okay, cool. Now let's see what percentage of cells that are doublets were also cells not included in the filtered nuclei.

```{r}
# find barcodes filtered out through high thresholds
nuclei_high_count <- nuclei_unfiltered |> 
  subset(
    tscp_count >= 100000 &
    gene_count >= 8000
  )
bc_high_count <- rownames(nuclei_high_count@meta.data)

# find barcodes in unfiltered but not filtered
bc_unfilt_not_filt <- setdiff(bc_unfiltered, bc_filtered)
```

```{r plot overlaps}
bc.list <- list(Filtered = bc_filtered,
                Unfiltered = bc_unfiltered)
venndir(setlist = bc.list, 
        proportional = T,
        show_labels = 'NPc',
        main = 'Filtered and Unfiltered May2024 data:\n(Unfiltered nuclei include high-count nuclei)')


bc.list <- list(In_Unfiltered_But_Not_Filtered = bc_unfilt_not_filt,
                High_Count_Nuclei = bc_high_count)
venndir(setlist = bc.list, 
        proportional = T,
        show_labels = 'NPc',
        main = 'Filtered and Unfiltered May2024 data:\n(Unfiltered nuclei include high-count nuclei)')


bc.list <- list(Doublets = bc_doublets,
                Filtered_nuclei = bc_filtered)
venndir(setlist = bc.list, 
        proportional = T,
        show_labels = 'NPc',
        main = 'Most detected doublets remain in the "filtered" dataset')


bc.list <- list(Doublets = bc_doublets,
                High_Count_Nuclei = bc_high_count)
venndir(setlist = bc.list, 
        proportional = T,
        show_labels = 'NPc',
        main = 'Most detected doublets are not high-count nuclei')
```
