---
title: "UMAPs of combined subclasses using union circadian DEGs"
format:
  html:
    code-fold: true
    page-layout: full
    toc: true
    toc-location: right
execute:
  warning: false
  message: false
---

The UMAPs produced here are constructed of:

-   A gene space defined by the union of circadian DEGs in each of the subclasses
-   All nuclei from each of the target subclasses (CA1, DG, Astro, Oligo)

```{r}
#| label: load-libs-data

source('03-scripts/R/seq_functions.R')

# ---- load data ----
nuclei <- LoadDataset('Dec2024')

activity_colors <- LoadActivityColors('Dec2024')
subclass_colors <- LoadAllenColors('subclass')
zt_colors <- LoadZTColors()
```


```{r}
#| label: prep-plots

# ---- per-subclass circadian DEGs (ZT in SE) ----
deg_dir <- '04-analysis/DEGs/Dec2024_ZT_SE_pseudobulk'

target_subclasses <- c(
  '016 CA1-ProS Glut',
  '037 DG Glut',
  '319 Astro-TE NN',
  '327 Oligo NN'
)

# map subclass names to circadian DEG file stems
group_stem <- c(
  '016 CA1-ProS Glut' = 'CA1',
  '037 DG Glut'       = 'DG',
  '319 Astro-TE NN'   = 'Astro',
  '327 Oligo NN'      = 'Oligo'
)
deg_files <- setNames(file.path(deg_dir, paste0(unname(group_stem[names(group_stem)]), '_all_DEGs.csv')),
                      names(group_stem))

deg_genes_by_subclass <- lapply(names(deg_files), function(s) {
  f <- deg_files[[s]]
  stopifnot(file.exists(f))
  read_csv(f, show_col_types = FALSE) |>
    filter(classification != 'no_change') |>
    pull(gene) |>
    unique()
})
names(deg_genes_by_subclass) <- names(deg_files)

# Build the union of circadian DEGs across all subclasses
union_degs <- sort(unique(unlist(deg_genes_by_subclass)))

# ---- single-object UMAP using union circadian DEGs ----
# Combine all target subclasses and restrict to SE (ZT is only defined in SE)
nuclei_combined <- nuclei |>
  subset(subclass_name %in% target_subclasses) |>
  subset(activity_condition == 'SE')

# Only keep DEGs present in this combined object
union_degs_in_data <- intersect(union_degs, rownames(nuclei_combined))

# Normalize, scale, and run reductions on shared circadian DEG space
nuclei_combined <- NormalizeData(nuclei_combined, verbose = FALSE)
nuclei_combined <- ScaleData(nuclei_combined, features = union_degs_in_data, verbose = FALSE)

n_pcs <- min(41, length(union_degs_in_data))
nuclei_combined <- RunPCA(
  nuclei_combined,
  features = union_degs_in_data,
  npcs = n_pcs,
  reduction.name = 'pca.deg',
  verbose = FALSE
)
nuclei_combined <- RunUMAP(
  nuclei_combined,
  reduction = 'pca.deg',
  dims = 1:n_pcs,
  reduction.name = 'umap.deg',
  verbose = FALSE,
  seed.use = 17
)
```

These first plots show UMAPs of all four subclasses together, colored by subclass and ZT.

```{r}
#| label: umap-1x2
#| fig-subcap: ["By subclass", "By ZT"]
#| layout-ncol: 2

# Prepare color vectors for present levels only
present_subclasses <- sort(unique(nuclei_combined$subclass_name))
present_zt <- sort(unique(nuclei_combined$ZT))
subclass_colors_present <- subclass_colors[present_subclasses]
zt_colors_present <- zt_colors[present_zt]

# Build plots
plots <- list()

p_by_subclass <- DimPlot(
  nuclei_combined,
  reduction = 'umap.deg',
  group.by = 'subclass_name',
  pt.size = 0.1,
  shuffle = TRUE,
  seed = 17,
  raster = FALSE
) +
  scale_color_manual(values = subclass_colors_present, drop = FALSE) +
  theme_minimal() +
  labs(title = 'Union circadian DEGs (all subclasses together) — colored by subclass', x = '', y = '') +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    # legend.position = 'none'
  )

p_by_zt <- DimPlot(
  nuclei_combined,
  reduction = 'umap.deg',
  group.by = 'ZT',
  pt.size = 0.1,
  shuffle = TRUE,
  seed = 17,
  raster = FALSE
) +
  scale_color_manual(values = zt_colors_present, drop = FALSE) +
  theme_minimal() +
  labs(title = 'Union circadian DEGs (all subclasses together) — colored by ZT', x = '', y = '') +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    # legend.position = 'none'
  )

print(p_by_subclass)
print(p_by_zt)
plots[['by_subclass']] <- p_by_subclass
plots[['by_ZT']] <- p_by_zt

# print interactive versions of UMAPs
# HoverLocator(p_by_zt, information = nuclei_combined@meta.data)
```

These next plots show the same UMAPs, with only one subclass highlighted at a time. The subclass nuclei are colored by ZT.

```{r}
#| label: umap-2x2
#| fig-subcap: ["CA1", "DG", "Astro", "Oligo"]
#| layout-ncol: 2
#| out-width: 100%

# ---- subclass-highlighted UMAPs (color by ZT for one subclass; others grey) ----
# Approach: create a masked metadata column where non-target cells are labeled 'Other'.
for (sub in target_subclasses) {
  # Mask ZT for non-target subclasses
  nuclei_combined$ZT_masked <- ifelse(
    nuclei_combined$subclass_name == sub,
    as.character(nuclei_combined$ZT),
    'Other'
  )
  nuclei_combined$ZT_masked <- factor(nuclei_combined$ZT_masked)

  # Build color vector: grey for 'Other' + ZT palette for present levels
  present_lvls <- levels(nuclei_combined$ZT_masked)
  zt_lvls <- intersect(names(zt_colors), present_lvls)
  color_map <- c('Other' = 'grey80')
  if (length(zt_lvls) > 0) {
    color_map <- c(color_map, zt_colors[zt_lvls])
  }

  p_masked <- DimPlot(
    nuclei_combined,
    reduction = 'umap.deg',
    group.by = 'ZT_masked',
    pt.size = 0.1,
    shuffle = TRUE,
    seed = 17,
    raster = FALSE
  ) +
    scale_color_manual(values = color_map, drop = FALSE) +
    theme_minimal() +
    labs(title = glue('{sub} highlighted (ZT); others grey'), x = '', y = '') +
    theme(
      panel.grid = element_blank(),
      axis.text = element_blank(),
    #   legend.position = 'none'
    )

  print(p_masked)
  plots[[glue('highlight__{sub}')]] <- p_masked
}
```