---
title: "Reduced UMAP workbook"
author: "jack"
editor_options: 
  chunk_output_type: console
---

```{r load everything}
library(ggplot2)
library(dplyr)
library(readr)
library(glue)
library(patchwork)

library(Seurat)

source("03-scripts/R/seq_functions.R")
activity_colors <- LoadActivityColors("Dec2024")
subclass_colors <- LoadAllenColors("subclass")
supertype_colors <- LoadAllenColors("supertype")


# establish subclasses to use ----------------------------------------------------
print('Subsetting Seurat object...')
gigaclasses <- LoadSubclassesToUse(nuclei, as_gigaclasses = T)
seurat_subsets <- LoadDataset("Dec2024", as_gigaclasses = T)
```

Make UMAPs plotting gigaclasses in IEG space

```{r}
# load DEGs ---------------------------------------------------------------------
print("Getting DEGs for all subclasses and contrasts...")
gene_list <- LoadGeneList('DEGs')
gene_list <- LoadGeneList("IEG")

# re-embed cells in this gene space -----------------------------------
for (gigaclass in names(gigaclasses)) {
  print(glue("Creating subsetted Seurat object for {gigaclass}..."))
  # re-embed cells in DEG space ------------------------------------------------------------
  print(glue("Re-embedding {gigaclass} cells in DEG space..."))
  nuclei_tmp_processed <- seurat_subsets[[gigaclass]] |> 
    subset(activity_condition %in% c('SE', 'EE30m', 'EE6h')) |> 
    RunPCA(features = gene_list) |> 
    FindNeighbors(dims = 1:9) |> 
    FindClusters(resolution = 0.8, random.seed = 17) |> 
    RunUMAP(dims = 1:9, seed.use = 17)
  
  
  # plot reduced UMAP ------------------------------------------------------------
  p1 <- DimPlot(nuclei_tmp_processed, reduction = 'umap', group.by = 'subclass_name') +
    scale_color_manual(values = subclass_colors) +
    labs(title = glue("{gigaclass} re-embed")) +
    theme_void() +
    theme(legend.position = 'none')
  
  p2 <- DimPlot(nuclei_tmp_processed, reduction = 'umap', group.by = 'activity_condition') +
    scale_color_manual(values = activity_colors) +
    labs(title = glue("{gigaclass} re-embed")) +
    theme_void() +
    theme(legend.position = 'none')
  
  p3 <- DimPlot(nuclei_tmp_processed, reduction = 'umap') +
    labs(title = glue("{gigaclass} re-embed")) +
    theme_void()
  
  print(p1 + p2 + p3)
  
  plots <- list()
  for (gene in gene_list) {
    print(glue("Plotting {gene}..."))
    
    p1 <- FeaturePlot(nuclei_tmp_processed, features = gene) +
      labs(title = glue("{gigaclass} {gene}")) +
      theme_void() +
      theme(legend.position = 'none')
    
    plots[[length(plots) + 1]] <- p1
  }
  p_mega <- wrap_plots(plots, ncol = 3, nrow = 5)
  print(p_mega)
}
```

Make UMAPs plotting gigaclasses in matched gene space

```{r}
# load DEGs ---------------------------------------------------------------------
print("Getting DEGs for all subclasses and contrasts...")
gene_list <- matched_genes_excitatory # from figure3_findMatchedExpressionGenes

# re-embed cells in this gene space -----------------------------------
print(glue("Re-embedding {gigaclass} cells in matched-gene space..."))
for (gigaclass in c('inhibitory', 'glia')) {
  nuclei_tmp_processed <- seurat_subsets[[gigaclass]] |> 
    subset(activity_condition %in% c('SE', 'EE30m', 'EE6h')) |> 
    RunPCA(features = gene_list) |> 
    FindNeighbors(dims = 1:6) |> 
    FindClusters(resolution = 0.8, random.seed = 17) |> 
    RunUMAP(dims = 1:6, seed.use = 17)
  
  
  # plot reduced UMAP ------------------------------------------------------------
  p1 <- DimPlot(nuclei_tmp_processed, reduction = 'umap', group.by = 'subclass_name', pt.size = 2) +
    scale_color_manual(values = subclass_colors) +
    labs(title = glue("{gigaclass} re-embed")) +
    # theme_void() +
    theme(legend.position = 'none')
  
  p2 <- DimPlot(nuclei_tmp_processed, reduction = 'umap', group.by = 'activity_condition', pt.size = 2) +
    scale_color_manual(values = activity_colors) +
    labs(title = glue("{gigaclass} re-embed")) +
    theme_void() +
    theme(legend.position = 'none')
  
  p3 <- DimPlot(nuclei_tmp_processed, reduction = 'umap', pt.size = 2) +
    labs(title = glue("{gigaclass} re-embed")) +
    theme_void() +
    theme(legend.position = 'none')
  
  print(p1 + p2 + p3)
  
  plots <- list()
  for (gene in gene_list) {
    print(glue("Plotting {gene}..."))
    
    p1 <- FeaturePlot(nuclei_tmp_processed, reduction = 'umap', features = gene) +
      labs(title = glue("{gigaclass} {gene}")) +
      theme_void() +
      theme(legend.position = 'none')
    
    plots[[length(plots) + 1]] <- p1
  }
  p_mega <- wrap_plots(plots, ncol = 3, nrow = 5)
  print(p_mega)
}
```

The overall shape of the UMAPs look very strange. I wonder if this is a function of embedding in a relatively low number of dimensions (in this case 9 for the IEGs).

Let me try embedding in all-DEG space but using only 9 PCs for the neighbor search and UMAP.

```{r}
# load DEGs ---------------------------------------------------------------------
print("Getting DEGs for all subclasses and contrasts...")
gene_list <- LoadGeneList('DEGs')
ieg_symbols <- LoadGeneList('IEG')

# re-embed cells in this smaller gene space ------------------------------------------------------------
for (gigaclass in names(gigaclasses)) {
  print(glue("Creating subsetted Seurat object for {gigaclass}..."))
  print(glue("Re-embedding {gigaclass} cells in DEG space..."))
  nuclei_tmp_processed <- seurat_subsets[[gigaclass]] |> 
    subset(activity_condition %in% c('SE', 'EE30m', 'EE6h')) |> 
    RunPCA(features = gene_list) |> 
    FindNeighbors(dims = 1:9) |> 
    FindClusters(resolution = 0.8, random.seed = 17) |> 
    RunUMAP(dims = 1:9, seed.use = 17)
  
  
  # plot reduced UMAP ------------------------------------------------------------
  p1 <- DimPlot(nuclei_tmp_processed, reduction = 'umap', group.by = 'cluster_name', pt.size = 2) +
    scale_color_manual(values = LoadAllenColors('cluster')) +
    labs(title = glue("{gigaclass} re-embed")) +
    theme_void() +
    theme(legend.position = 'none')
  
  p2 <- DimPlot(nuclei_tmp_processed, reduction = 'umap', group.by = 'activity_condition') +
    scale_color_manual(values = activity_colors) +
    labs(title = glue("{gigaclass} re-embed")) +
    theme_void() +
    theme(legend.position = 'none')
  
  p3 <- DimPlot(nuclei_tmp_processed, reduction = 'umap') +
    labs(title = glue("{gigaclass} re-embed")) +
    theme_void()
  
  print(p1 + p2 + p3)
  
  plots <- list()
  for (gene in ieg_symbols) {
    print(glue("Plotting {gene}..."))
    
    p1 <- FeaturePlot(nuclei_tmp_processed, features = gene) +
      labs(title = glue("{gigaclass} {gene}")) +
      theme_void() +
      theme(legend.position = 'none')
    
    plots[[length(plots) + 1]] <- p1
  }
  p_mega <- wrap_plots(plots, ncol = 3, nrow = 5)
  print(p_mega)
}
```

Plot in glial ERG space

```{r}
df_expression <- read_csv(glue("04-analysis/df_expression/df_expression.seurat/df_expression_glia.csv"))
gene_list <- df_expression |> 
  filter(classification == 'ERG') |> 
  pull(gene)

print(glue("Creating subsetted Seurat object for {gigaclass}..."))
nuclei_tmp_processed <- seurat_subsets[['glia']] |> 
  subset(activity_condition %in% c('SE', 'EE30m', 'EE6h')) |> 
  RunPCA(features = gene_list) |> 
  FindNeighbors(dims = 1:26) |> 
  FindClusters(resolution = 0.8, random.seed = 17) |> 
  RunUMAP(dims = 1:26, seed.use = 17)


# plot reduced UMAP ------------------------------------------------------------
p1 <- DimPlot(nuclei_tmp_processed, reduction = 'umap', group.by = 'subclass_name', pt.size = 2) +
  scale_color_manual(values = LoadAllenColors('subclass')) +
  labs(title = glue("{gigaclass} re-embed")) +
  theme_void() +
  theme(legend.position = 'none')

p2 <- DimPlot(nuclei_tmp_processed, reduction = 'umap', group.by = 'supertype_name', pt.size = 2) +
  scale_color_manual(values = LoadAllenColors('supertype')) +
  labs(title = glue("{gigaclass} re-embed")) +
  theme_void() +
  theme(legend.position = 'none')

p3 <- DimPlot(nuclei_tmp_processed, reduction = 'umap', group.by = 'cluster_name', pt.size = 2) +
  scale_color_manual(values = LoadAllenColors('cluster')) +
  labs(title = glue("{gigaclass} re-embed")) +
  theme_void() +
  theme(legend.position = 'none')

p4 <- DimPlot(nuclei_tmp_processed, reduction = 'umap', group.by = 'activity_condition') +
  scale_color_manual(values = activity_colors) +
  labs(title = glue("{gigaclass} re-embed")) +
  theme_void() +
  theme(legend.position = 'none')

p5 <- DimPlot(nuclei_tmp_processed, reduction = 'umap') +
  labs(title = glue("{gigaclass} re-embed")) +
  theme_void()

print(p1 + p2 + p3 + p4 + p5)

plots <- list()
for (gene in ieg_symbols) {
  print(glue("Plotting {gene}..."))
  
  p1 <- FeaturePlot(nuclei_tmp_processed, features = gene) +
    labs(title = glue("{gigaclass} {gene}")) +
    theme_void() +
    theme(legend.position = 'none')
  
  plots[[length(plots) + 1]] <- p1
}
p_mega <- wrap_plots(plots, ncol = 3, nrow = 5)
print(p_mega)
```

Let's make a UMAP in DEG space (50 PCs) and then plot high-tau genes on that UMAP

```{r}
gene_list <- LoadGeneList('DEGs')
df_expression <- read_csv(glue("04-analysis/df_expression/df_expression.seurat/df_expression_excitatory.csv"))
  
df_sub <- df_expression |>
    filter(activity_condition == 'EE30m',
           classification == 'ERG',
           direction == 'up'
           ) |>
    group_by(gene) |> 
    mutate(tau = tau(log2FoldChange.shrink, direction = 'up')) |>
    select(gene, subclass, log2FoldChange.shrink, tau)
  
genes_to_plot_high_tau <- df_sub |> 
  arrange(desc(tau)) |> 
  distinct(gene) |> 
  head(10) |> 
  pull() |> 
  print()
  
nuclei_tmp_processed <- seurat_subsets[['excitatory']] |> 
  subset(activity_condition %in% c('SE', 'EE30m', 'EE6h')) |> 
  RunPCA(features = gene_list) |> 
  FindNeighbors(dims = 1:50) |> 
  FindClusters(resolution = 0.8, random.seed = 17) |> 
  RunUMAP(dims = 1:50, seed.use = 17)


# plot reduced UMAP ------------------------------------------------------------
p1 <- DimPlot(nuclei_tmp_processed, reduction = 'umap', group.by = 'subclass_name', pt.size = 2) +
  scale_color_manual(values = LoadAllenColors('subclass')) +
  labs(title = glue("{gigaclass} re-embed")) +
  theme_void() +
  theme(legend.position = 'none')

p2 <- DimPlot(nuclei_tmp_processed, reduction = 'umap', group.by = 'supertype_name', pt.size = 2) +
  scale_color_manual(values = LoadAllenColors('supertype')) +
  labs(title = glue("{gigaclass} re-embed")) +
  theme_void() +
  theme(legend.position = 'none')

p3 <- DimPlot(nuclei_tmp_processed, reduction = 'umap', group.by = 'cluster_name', pt.size = 2) +
  scale_color_manual(values = LoadAllenColors('cluster')) +
  labs(title = glue("{gigaclass} re-embed")) +
  theme_void() +
  theme(legend.position = 'none')

p4 <- DimPlot(nuclei_tmp_processed, reduction = 'umap', group.by = 'activity_condition') +
  scale_color_manual(values = activity_colors) +
  labs(title = glue("{gigaclass} re-embed")) +
  theme_void() +
  theme(legend.position = 'none')

p5 <- DimPlot(nuclei_tmp_processed, reduction = 'umap') +
  labs(title = glue("{gigaclass} re-embed")) +
  theme_void() +
  theme(legend.position = 'none')

print(p1 + p2 + p3 + p4 + p5)

plots <- list()
for (gene in genes_to_plot_high_tau) {
  print(glue("Plotting {gene}..."))
  
  p1 <- FeaturePlot(nuclei_tmp_processed, features = gene) +
    labs(title = glue("{gigaclass} {gene}")) +
    theme_void() +
    theme(legend.position = 'none')
  
  plots[[length(plots) + 1]] <- p1
}
p_mega <- wrap_plots(plots, ncol = 5, nrow = 2)
print(p_mega)

VlnPlot(nuclei_tmp_processed, group.by = 'subclass_name', split.by = 'activity_condition')
```
