---
title: "Figure 1 analysis"
author: "jack"
editor_options: 
  chunk_output_type: console
---

# load in data

```{r}
library(Seurat)
library(SeuratDisk)

library(tidyverse)
library(readxl)

library(paletteer)
library(hues)

library(ComplexHeatmap)
library(dendextend)
library(circlize)
library(glue)
library(patchwork)
library(ggtext)
library(scales)
```

# init

```{r}
# load in all custom functions
source("~/code/seq/manuscript_code/seq_functions.R")

nuclei <- LoadDataset(dataset = 'May2024', sublibrary = 'combined')
```

TF list taken from [AnimalTFDBv4.0](http://bioinfo.life.hust.edu.cn/AnimalTFDB4/#/) (Shen et al. (2023). *Nuc. Acids Res.*). [TFcheckpoint](https://www.tfcheckpoint.org/index.php/data-sources) is also a good resource.

```{r}
palette_colors <- paletteer_d('ggthemes::wsj_colors6')
custom_colors <- c('dSE' = palette_colors[2], 
                   'd30m' = palette_colors[1],
                   'd6h' = palette_colors[3],
                   'KA' = palette_colors[4])

tf <- readxl::read_excel("/Users/jack/code/seq/data/Mus_musculus_TF.xlsx")

# get genes of interest
dSE_vs_d30m <- readxl::read_excel("~/code/seq/data/DEGs/016_CA1-ProS_Glut_d30m_vs_dSE.csv")
dSE_vs_d6h <- readxl::read_excel()
dSE_vs_KA <- readxl::read_excel("~/code/seq/data/DEGs/016_CA1-ProS_Glut_KA_vs_dSE.csv")
d30m_vs_d6h <- readxl::read_excel()

ieg_symbols <- c(
  'Arc',
  'Btg2',
  'Dusp5',
  'Dusp1',
  'Egr1',
  'Egr2',
  'Egr3',
  'Egr4',
  'Fosb',
  'Fos',
  'Junb',
  'Npas4',
  'Nr4a1',
  'Ppp1r15a',
  'Map3k19',
  'Bdnf'
)
```

# gene ComplexHMs

correlation matrices for complex heatmaps

```{r}
degs  <- '___dSE-vs-d30m-DEGs'
gene_list <- dSE_vs_d30m |> 
  # filter(gene %in% tf$Symbol) |>
  filter(p_val_adj_MAST <= 0.01) |>
  filter(abs(avg_log2FC) > 0.5) |>
  pull(gene)

# define active cells
df_active_cells <- FindActiveCells(nuclei)
active_cells_seurat <- nuclei |> subset(bc_wells %in% df_active_cells$cell)
ca1_seurat <- nuclei |> subset(subclass_name == '016 CA1-ProS Glut')

corr_dSE_active  <- GetCorrData(active_cells_seurat, 'dSE', gene_list)
corr_d30m_active <- GetCorrData(active_cells_seurat, 'd30m', gene_list)
corr_d6h_active  <- GetCorrData(active_cells_seurat, 'd6h', gene_list)
corr_KA_active   <- GetCorrData(active_cells_seurat, 'KA', gene_list)

corr_dSE_all     <- GetCorrData(ca1_seurat,          'dSE', gene_list)
corr_d30m_all    <- GetCorrData(ca1_seurat,          'd30m', gene_list)
corr_d6h_all     <- GetCorrData(ca1_seurat,          'd6h', gene_list)
corr_KA_all      <- GetCorrData(ca1_seurat,          'KA', gene_list)

# plot the heatmaps
PlotComplexHeatmap(corr_dSE_all,     paste0('dSE-all', degs), save_plot = TRUE)
PlotComplexHeatmap(corr_dSE_active,  paste0('dSE-active', degs), save_plot = TRUE)
PlotComplexHeatmap(corr_d30m_all,    paste0('d30m-all', degs), save_plot = TRUE)
PlotComplexHeatmap(corr_d30m_active, paste0('d30m-active', degs), save_plot = TRUE)
PlotComplexHeatmap(corr_d6h_all,     paste0('d6h-all', degs), save_plot = TRUE)
PlotComplexHeatmap(corr_d6h_active,  paste0('d6h-active', degs), save_plot = TRUE)
PlotComplexHeatmap(corr_KA_all,      paste0('KA-all', degs), save_plot = TRUE)
PlotComplexHeatmap(corr_KA_active,   paste0('KA-active', degs), save_plot = TRUE)
```

# ITF correlation matricies

I want to see if there are any TFs that are co-activated together. This will result in an NxN correlation matrix of co-expression for transcription factors that are induced at either 30m or KA compared to HC.

Note: THIS IS ONLY FOR CA1 CELLS RIGHT NOW

```{r}
# get TFs genome-wide in the mouse
tf_list <- "~/code/seq/data/Mus_musculus_TF.xlsx" |> 
  readxl::read_excel() |> 
  pull(Symbol)

dSE_vs_d30m_list <- dSE_vs_d30m |> pull(gene)
dSE_vs_KA_list <- dSE_vs_KA |> pull(gene)
induced_genes <- c(dSE_vs_d30m_list)

itfs <- tf_list[tf_list %in% induced_genes]
```

```{r}
# get right cell subsets
ca1_seurat <- nuclei |> subset(subclass_name == '016 CA1-ProS Glut')

df_active_cells <- FindActiveCells(nuclei) |> 
  filter(active_binary == TRUE)
ca1_seurat_active <- nuclei |> 
  subset(bc_wells %in% df_active_cells$cell) |> 
  subset(subclass_name == '016 CA1-ProS Glut')

# get the correlation matrix for the TFs
corr_d30m_active <- GetCorrData(ca1_seurat_active, 'd30m', itfs, output_fmt = 'complex_heatmap')
corr_d30m_all    <- GetCorrData(ca1_seurat,        'd30m', itfs, output_fmt = 'complex_heatmap')

# plot correlation matricies
source("~/code/seq/manuscript_code/seq_functions.R")
PlotComplexHeatmap(corr_d30m_active, 'd30m-active-TFs', save_plot = FALSE)
PlotComplexHeatmap(corr_d30m_all, 'd30m-all-TFs', save_plot = FALSE)
```

This cell creates some scatter plots comparing Foxo6 expression to the expression of other ITFs. It also uses Egr2 as a comparison ITF with low copy number that is positively correlated with other canonical ITFs.

```{r}
# create df of expression for itfs
df_expression <- ca1_seurat |> 
  subset(condition == 'd6h') |>
  GetAssayData('SCT', layer = 'data') |> 
  t() |>
  as.data.frame() |> 
  rownames_to_column(var = 'cell')

df_expression |> 
  pivot_longer(cols = -c(cell, Foxo6), names_to = 'gene', values_to = 'expression') |>
  filter(gene %in% itfs) |> 
ggplot() +
  aes(x = Foxo6, y = expression) +
  geom_jitter(alpha = 0.3) +
  facet_wrap(~gene)

df_expression |> 
  pivot_longer(cols = -c(cell, Egr2), names_to = 'gene', values_to = 'expression') |>
  filter(gene %in% itfs) |> 
ggplot() +
  aes(x = Egr2, y = expression) +
  geom_jitter(alpha = 0.3) +
  facet_wrap(~gene)

###### code for meeting with brenda
# df_expression |> 
#   pivot_longer(cols = -c(cell, Npas4), names_to = 'gene', values_to = 'expression') |>
#   filter(gene %in% itfs) |> 
# ggplot() +
#   aes(x = Npas4) +
#   geom_jitter(alpha = 0.3) +
#   facet_wrap(~gene)
```

# compare the expression distribution of ITFs

```{r}
df_expression |> 
  pivot_longer(cols = -c(cell), names_to = 'gene', values_to = 'expression') |>
  filter(gene %in% itfs) |> 
ggplot() +
  geom_histogram(aes(x = expression)) +
  facet_wrap(~gene, scales = 'free')

```

# hclust

I want to see if there are genes that form clusters of expression among the IEGs. I will use hierarchical clustering to do this.

I would like to distinguish between active and inactive cells when doing this, so I'm going to take some code that I used in my figure 1 workbook and paste it here.

```{r}
df_active_cells <- FindActiveCells(nuclei)

# df_expression has metadata and expression info. metadata cols start with 'm_'
nuclei_meta <- nuclei@meta.data |> as_tibble() |>
  select(starts_with('supertype') | 'bc_wells') |> 
  rename(m_cell = bc_wells) |>
  relocate(m_cell) |> 
  # rename cols that start with 'supertype' to start with 'm_supertype'
  rename_with(~str_replace(., '^supertype', 'm_supertype'), starts_with('supertype'))
    
df_expression <- nuclei |> 
  subset(subclass_name == '016 CA1-ProS Glut') |>
  GetAssayData('SCT', layer = 'scale.data') |> 
  as.data.frame() |> 
  rownames_to_column(var = 'gene') |>
  filter(gene %in% gene_list) |> 
  pivot_longer(cols = -gene, names_to = 'm_cell', values_to = 'expression') |> 
  pivot_wider(names_from = gene, values_from = expression) |> 
  left_join(df_active_cells) |> 
  select(-cell) |> 
  relocate(condition, active_binary, num_upregd_IEGs, .after = m_cell) |> 
  rename(m_condition = condition, m_active_binary = active_binary, m_num_upregd_IEGs = num_upregd_IEGs) |> 
  left_join(nuclei_meta) |> 
  # select cols that start with m_ or whose colnames are in gene_list
  select(starts_with('m_') | any_of(gene_list)) |>
  relocate(starts_with('m_supertype'), .after = m_num_upregd_IEGs)
```

## active cells

```{r}
condition <- 'd30m'
hclust_gene_mat_active <- df_expression |>
  filter(m_condition == condition) |>
  filter(m_active_binary == TRUE) |>
  column_to_rownames(var = 'm_cell') |>
  select(-starts_with('m_')) |> # remove metadata cols to force numeric
  # remove cols with all 0s
  select(where(~!all(. == 0))) |> 
  # re-center column values to the range of expression for these cells
  mutate(across(!starts_with('m_'), ~(. - mean(.)) / sd(.))) |> 
  as.matrix()

# perform hclustering on scaled gene data
dend_active_cells <- hclust_gene_mat_active |> 
  scale() |> 
  t() |> 
  dist() |> 
  hclust(method = 'complete') |>
  as.dendrogram() |> 
  # color_branches(k=6, col = paletteer_d('ggthemes::wsj_colors6')) |>
  # color_labels(k=6, col = paletteer_d('ggthemes::wsj_colors6')) |>
  set('branches_lwd', 2) |> 
  set('labels_cex', 1)

plot(dend_active_cells)
title(paste0('Active cells CA1 ', condition))
```

## all cells

```{r}
hclust_gene_mat_all <- df_expression |>
  filter(m_condition == condition) |>
  column_to_rownames(var = 'm_cell') |>
  select(-starts_with('m_')) |> # remove metadata cols to force numeric
  # remove cols with all 0s
  select(where(~!all(. == 0))) |> 
  as.matrix()

# perform hclustering on scaled gene data
dend_all_cells <- hclust_gene_mat_all |> 
  scale() |> 
  t() |> 
  dist() |> 
  hclust(method = 'complete') |>
  as.dendrogram() |> 
  # color_branches(k=6, col = paletteer_d('ggthemes::wsj_colors6')) |>
  # color_labels(k=6, col = paletteer_d('ggthemes::wsj_colors6')) |>
  set('branches_lwd', 2) |> 
  set('labels_cex', 0.8)

plot(dend_all_cells)
title(paste0('All cells CA1 ', condition))
```

# cell heatmap

I want to try my own heatmap implementation, instead of Seurat's shitty one.

```{r}
# get module assignment and then map module number to a color
k_clusters = 6
df_module_assignments <- hclust_gene_mat_active |>
  scale() |>
  t() |>
  dist() |>
  hclust(method = 'complete') |>
  cutree(k_clusters) |> 
  as.data.frame() |> 
  rownames_to_column(var = 'gene') |> 
  rename(gene_module = 2) |>
  mutate(gene = factor(gene, levels = labels(dend_active_cells))) |> 
  as_tibble() |> 
  arrange(gene) |> 
  mutate(color = paletteer_d('ggthemes::wsj_colors6')[gene_module]) |> 
  mutate(gene_str = glue("<i style='color:{color}'>{gene}</i>")) |> print()

# make df for plotting
df_gg <- df_expression |> 
  filter(m_condition == condition) |>
  pivot_longer(cols = -starts_with('m_'), names_to = 'gene', values_to = 'expression') |>
  left_join(df_module_assignments) # incorporate gene module info

order_by_module <- df_gg |>
  summarize(module_avg_expression = mean(expression), .by = c(m_cell, gene_module)) |>
  # make new variable which denotes which cluster for each cell has the highest average expression
  mutate(highest_module = which(module_avg_expression == max(module_avg_expression)), .by = m_cell) |> 
  # arrange(m_supertype_name, -m_supertype_bootstrapping_probability) |> 
  arrange(highest_module, module_avg_expression) |> 
  filter(gene_module == highest_module)

p_grid <- df_gg |> 
  left_join(order_by_module) |> 
ggplot() +
  aes(x = gene_str, y = m_cell, fill = expression) +
  geom_tile() +
  scale_y_discrete(limits = unique(order_by_module$m_cell)) +
  scale_fill_gradient2(low = '#075AFF', mid = 'white', high = '#FF0000', midpoint = 0, limits = c(-2.5, 2.5), oob = squish) +
theme_minimal() +
  theme(
    axis.text.x = element_markdown(size = 10, angle = 90, hjust = 1, valign = 1),
    axis.text.y = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    # legend.position = 'none',
  )

# create an annotation bar to go along the y-axis that shows the cell type
p_bar <- order_by_module |> 
  left_join(df_module_assignments) |>
ggplot() +
  geom_tile(aes(x = 1, y = m_cell, fill = color)) +
  scale_y_discrete(limits = order_by_module$m_cell) +
  # scale_fill_discrete(paletteer_d('ggthemes::wsj_colors6')) +
  scale_fill_identity() +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    plot.margin  = margin(r = 0, l = 0),
    legend.position = 'none'
  )

p_bar + p_grid + p_bar + plot_layout(widths = c(1, 100, 1))
```
